
## 14. Across communities, are estimates biased?

The September fits can answer more questions without being fitted again. We now use the same saved results to distinguish **bias**, **typical error** and **interval coverage**. All four packages remain in this comparison. These calculations extend the ten-community experiment; they do not add new simulated communities.

An estimate's **signed error** is its estimate minus truth. Averaging signed errors over independent simulated communities estimates bias: persistent overestimation is positive, and persistent underestimation is negative. Errors in opposite directions can cancel, so we also report **root mean squared error (RMSE)**. RMSE squares each error before averaging and then takes a square root. It describes error size and gives larger errors more weight than mean absolute error does.

```{r calibration-data}
calibration <- readRDS("teaching-data/lesson-4-calibration.rds")
stopifnot(calibration$complete)

prediction_calibration <- as_tibble(calibration$prediction) |>
  mutate(package = factor(package, levels = package_order))

prediction_summary <- prediction_calibration |>
  group_by(package, scenario, n_sites, n_species, response, use_traits) |>
  summarise(
    communities = n(),
    flagged = sum(!diagnostic_pass),
    bias_pp = 100 * mean(bias),
    bias_mcse_pp = if (n() > 1) 100 * sd(bias) / sqrt(n()) else NA_real_,
    rmse_pp = 100 * sqrt(mean(rmse^2)),
    .groups = "drop"
  )

prediction_summary |>
  filter(scenario == "baseline") |>
  select(package, n_sites, communities, flagged, bias_pp, bias_mcse_pp, rmse_pp) |>
  knitr::kable(digits = 2, col.names = c(
    "Package", "Sites", "Communities", "Flagged",
    "Bias (points)", "Bias MCSE (points)", "RMSE (points)"
  ))
```

Here the target is each species' occurrence probability at the 300 independent test sites, averaged over unknown hidden conditions. Each community receives equal weight. The **Monte Carlo standard error (MCSE)** describes how much the estimated average bias varies because we simulated only ten communities. It is not a species' posterior standard error. The thousands of species-by-site cells do not turn this into thousands of independent repetitions. RMSE is the square root of the average community mean squared error, not the average of community RMSEs.

```{r calibration-error-comparison, fig.height=8}
prediction_calibration |>
  filter(scenario != "traits") |>
  mutate(
    condition = paste(scenario, response, sep = ": "),
    diagnostic = if_else(diagnostic_pass, "Passed checks", "Unresolved"),
    sites = factor(n_sites)
  ) |>
  select(package, replicate, condition, diagnostic, sites, bias, rmse) |>
  pivot_longer(c(bias, rmse), names_to = "measure", values_to = "error") |>
  mutate(measure = recode(measure, bias = "Signed error", rmse = "RMSE")) |>
  ggplot(aes(package, 100 * error, colour = sites, shape = diagnostic)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(position = position_jitterdodge(jitter.width = .15, dodge.width = .6, seed = 25), alpha = .75) +
  facet_grid(condition ~ measure, scales = "free_y") +
  labs(x = NULL, y = "Error (percentage points)", colour = "Training sites", shape = "Diagnostics",
    caption = "Each point is one community. Flagged fits remain visible; failed fits have no score.") +
  theme_bw() + theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Small signed errors alongside appreciable RMSE mean that errors cancel, not that individual estimates are precise. Compare the distributions of community results, rather than choosing a winner from a small difference between averages. The failure counts in section 12 still apply. Where a package failed on some communities, its average describes only the communities with a scored fit.

## 15. Do 95% intervals contain the truth?

**Coverage** asks how often an interval includes the generating truth across repeated datasets. For example, an interval from 0.2 to 0.5 covers a true occurrence probability of 0.4 but misses a truth of 0.7. A narrow interval can be confidently wrong; a very wide interval can cover the truth while saying little. We therefore show **coverage and interval width together**.

Bayesian credible intervals and approximate frequentist confidence intervals have different definitions. Repeated simulation lets us measure the coverage of either procedure. Bayesian intervals are not guaranteed to have exactly 95% coverage under the ecological truth distribution used here.

### Which saved results support this calculation?

- **All four packages:** point errors for test-site probabilities and five shared environmental settings.
- **occJSDM and Hmsc:** 95% equal-tailed intervals for marginal occurrence probabilities, calculated from every saved global parameter draw.
- **occJSDM:** intervals for environmental coefficients, where the fitted response contains the generating terms and the logit coefficient definitions match.
- **occJSDM and gllvm:** existing intervals for the measured trait effects on their matching logit scale.
- **gllvm and sjSDM probability intervals, and their ordinary environmental-coefficient intervals:** unavailable from the records saved for this experiment. Missing intervals are recorded as unavailable, not as zero coverage.

Hmsc fitted a probit model to logit-generated data. Its probabilities can be compared with the true probabilities, but its raw coefficients do not have the same numerical truth as the logit coefficients. For coefficient recovery we restrict the calculation to conditions where the fitted response includes the generating terms for every species. The straight-response fit to the curved scenario is therefore omitted from the coefficient tables as a whole, including the species whose generating responses remain straight. Its probability errors and coverage still include every species. This study measures performance under the existing logit-generating design. A probit-generating arm is still needed for a balanced comparison of that modelling choice.

### Probability intervals at five shared environments

To keep the interval calculation inspectable, we evaluate five fixed settings: both environmental gradients at zero, then each gradient at -1 and +1 while the other remains zero. These are original simulation units, before training-data standardisation. Every package and community uses the same settings. Hidden site conditions are integrated out **within each posterior draw**, and the interval is then calculated across those marginal-probability draws. These are intervals for the average probability given the measured environment, not intervals for a particular site's hidden conditions or for a future binary observation.

```{r calibration-grid}
knitr::kable(calibration$grid, row.names = TRUE)
```

The probability coverage calculation therefore uses a different evaluation set from the 300-site point-error table above. Its mean, lower endpoint and upper endpoint were checked for numerical integration accuracy. It uses every archived draw, including all four chains. The original fitting diagnostics remain attached; post-processing cannot repair an unconverged chain.

```{r calibration-probability-summary}
probability_summary <- as_tibble(calibration$summary) |>
  filter(target == "Marginal probability", scenario == "baseline", population == "All scored fits") |>
  mutate(package = factor(package, levels = package_order)) |>
  arrange(package, n_sites)

probability_summary |>
  transmute(
    Package = package, Sites = n_sites, Communities = communities,
    Flagged = flagged_communities, `Communities with intervals` = interval_communities,
    `Coverage (%)` = 100 * coverage, `Coverage MCSE (points)` = 100 * coverage_mcse,
    `Mean width (points)` = 100 * width
  ) |>
  knitr::kable(digits = 1)
```

`NA` means that this archive cannot supply the interval comparison. Within each community we average coverage and width over its species and the five settings; we then average those community summaries equally. Coverage MCSE is calculated from the variation among the community summaries. It is not a binomial standard error that treats every species-setting pair as independent. With ten communities, the result is exploratory and the MCSE itself is uncertain.

```{r calibration-probability-intervals, fig.height=8}
as_tibble(calibration$per_community) |>
  filter(target == "Marginal probability", scenario != "traits", intervals > 0) |>
  mutate(
    condition = paste(scenario, response, sep = ": "),
    diagnostic = if_else(diagnostic_pass, "Passed checks", "Unresolved"),
    sites = factor(n_sites)
  ) |>
  select(package, replicate, condition, diagnostic, sites, coverage, width) |>
  pivot_longer(c(coverage, width), names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure, coverage = "Coverage (%)", width = "Mean width (points)")) |>
  ggplot(aes(package, 100 * value, colour = sites, shape = diagnostic)) +
  geom_hline(data = data.frame(measure = "Coverage (%)"), aes(yintercept = 95), linetype = 2, inherit.aes = FALSE) +
  geom_point(position = position_jitterdodge(jitter.width = .15, dodge.width = .6, seed = 25), alpha = .8) +
  facet_grid(condition ~ measure, scales = "free_y") +
  labs(x = NULL, y = NULL, colour = "Training sites", shape = "Diagnostics",
    caption = "Each point summarises one community. The dashed line marks nominal 95% coverage.") +
  theme_bw()
```

The straight-response fits to curved truth illustrate why width matters. At 300 sites, their measured probability coverage is `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "occJSDM" & scenario == "curved" & response == "linear" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` for occJSDM and `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "Hmsc" & scenario == "curved" & response == "linear" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` for Hmsc. Including the quadratic term gives `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "occJSDM" & scenario == "curved" & response == "quadratic" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` and `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "Hmsc" & scenario == "curved" & response == "quadratic" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)`, respectively. More data can narrow intervals around an unsuitable response shape; it does not supply the missing curve. These numbers describe the five settings and ten communities tested, including flagged fits.

The baseline is also not a substitute for checking difficult conditions. The rare-species and correlated-environment panels show lower occJSDM coverage than its baseline. This records a pattern to investigate with further replication and diagnostic work; it does not by itself identify a software defect or establish an overall package ranking.

All scored fits are retained above. A companion summary restricted to fits that passed the original diagnostics is also saved. Compare it with the full summary as a sensitivity check; excluding difficult fits can select easier communities, so the restricted summary is not a corrected package ranking.

```{r calibration-diagnostic-sensitivity}
as_tibble(calibration$summary) |>
  filter(target == "Marginal probability", scenario == "baseline", intervals > 0) |>
  transmute(Package = package, Sites = n_sites, Fits = population,
    Communities = interval_communities, `Coverage (%)` = 100 * coverage,
    `Mean width (points)` = 100 * width) |>
  knitr::kable(digits = 1)
```

### Environmental coefficients and trait relationships

For compatible coefficient targets we undo predictor scaling and centring. For the intercept, the adjustment is made jointly with each draw's slopes before taking quantiles. Rescaling a marginal intercept interval alone would discard the relevant dependence.

```{r calibration-environmental-effects}
as_tibble(calibration$summary) |>
  filter(target == "Environmental coefficient", scenario == "baseline", n_sites == 100,
    population == "All scored fits") |>
  transmute(Package = package, Coefficient = term, Communities = communities,
    Bias = bias, RMSE = rmse, `Coverage (%)` = 100 * coverage, `Mean width` = width) |>
  knitr::kable(digits = 3)
```

The coefficient table reports the same conditional ecological coefficient target as the simulator; it does not compare the orientation of latent factor axes. The probability comparison remains the shared target across all four packages. Hmsc is absent from this coefficient table because its link differs, not because it failed to supply estimates.

```{r calibration-trait-effects}
as_tibble(calibration$summary) |>
  filter(target == "Trait coefficient", n_sites == 100, n_species == 10,
    population == "All scored fits") |>
  transmute(Package = package, Relationship = term, Communities = communities,
    Flagged = flagged_communities, Bias = bias, `Coverage (%)` = 100 * coverage,
    `Mean width` = width) |>
  knitr::kable(digits = 3)
```

This displayed trait slice uses 100 sites and ten species; the saved summary also contains the 300-site and thirty-species cases. For the irrelevant trait, truth is zero, so an interval that misses truth also excludes zero incorrectly. For a genuinely nonzero effect, covering truth and excluding zero answer different questions. gllvm's original standard-error warnings and unavailable fits still matter; these are approximate intervals from the selected fits, not a guarantee of calibration.

The compact bundle contains per-community summaries and individual grid/coefficient records, so these tables can be filtered without accessing the large fitting archive. For example:

```{r calibration-inspect}
as_tibble(calibration$per_community) |>
  filter(package == "occJSDM", target == "Marginal probability",
    scenario == "baseline", n_sites == 100) |>
  select(replicate, diagnostic_pass, bias, rmse, coverage, width)
```

No model was refitted for these sections. A larger coverage study should add independent communities, complete the gllvm/sjSDM uncertainty methods, and include a probit-generating arm under a declared fitting and diagnostic protocol. Preserve these ten communities and their failures as the original experiment. The July/August occJSDM-only study predates model fixes and remains historical evidence; it is not pooled with these September results.
