
## 12. Does more data help when ecology becomes harder?

The original example above used one small community. The extension below repeats the experiment across **ten independently simulated communities**, so we can see how much the answer varies between datasets. Within each community, the 100 training sites are part of the 300 training sites. Both fits predict the same 300 independent test sites.

We change one ecological feature at a time. All four packages receive the same observations. The extension still assumes perfect observation and independent sites; field collection, PCR and spatial effects are not involved.

| Simulation | What changes | What we want to learn |
|---|---|---|
| Baseline | Neither of the complications below | Does collecting more sites improve estimates? |
| Rare species | Three species become much less common | Are their distributions especially difficult to estimate? |
| Correlated environment | The two predictors are strongly correlated | Can we distinguish their effects when they usually change together? |
| Curved responses | Five species prefer an intermediate value of gradient 1 | Does the fitted model need a curve to describe the relationship? |

### Which fits passed, remained flagged or failed?

The complete experiment has **640 package/dataset/model combinations**, before optimisation restarts: 400 ecological comparisons and 240 trait comparisons. Each counted fit represents one combination, even if fitting it required several starts or a longer MCMC run. All 640 were attempted. The table is calculated from the saved results, so unsuccessful combinations remain visible alongside the estimates we can score.

```{r extension-data}
extension <- readRDS("teaching-data/lesson-4-extension.rds")

extension_status <- as_tibble(extension$manifest) |>
  group_by(package) |>
  summarise(
    attempted = sum(completed),
    passed = sum(fit_ok & diagnostic_pass),
    flagged = sum(fit_ok & !diagnostic_pass),
    failed = sum(completed & !fit_ok),
    scored = sum(scored),
    .groups = "drop"
  ) |>
  arrange(match(package, c("occJSDM", "gllvm", "Hmsc", "sjSDM")))

extension_status <- bind_rows(
  extension_status,
  extension_status |>
    summarise(package = "Total", across(where(is.numeric), sum))
)

knitr::kable(
  extension_status,
  col.names = c("Package", "Attempted", "Passed", "Flagged", "Failed", "Scored")
)
```

**Run status:** `r if (extension$all_attempted) "All planned fitting combinations have been attempted. Check the diagnostic and scoring counts before interpreting the comparisons." else "The expanded study is still running. Results below are incomplete and must not be used to rank packages or draw conclusions about sample size."`

**Passed**, **flagged** and **failed** are mutually exclusive outcomes. A passed fit produced estimates and met the declared diagnostics. A flagged fit produced estimates that could be scored, but an unresolved diagnostic means we should treat them as provisional. A failed combination supplied no estimate accepted for scoring within the fixed fitting budget. Here, 505 passed plus 121 flagged fits give 626 scored fits; the remaining 14 failed combinations have no prediction-error score.

For **occJSDM and Hmsc**, we checked agreement among four MCMC chains and their effective sample sizes: Rhat at most 1.01 and bulk/tail ESS at least 400 for all monitored coefficients, residual covariances, fixed-grid probabilities and applicable trait effects. We also checked the numerical integration used for those probabilities. One longer fitting attempt was allowed if the initial checks failed. All 180 occJSDM fits passed the final checks in this experiment. The 86 flagged Hmsc fits still failed at least one Rhat/ESS criterion after the longer attempt; having posterior draws available did not make those diagnostics satisfactory.

For **gllvm and sjSDM**, we allowed up to six independent starts and asked whether another acceptable start reproduced the best solution: objective scores within 0.1 units and marginal probabilities within one percentage point on a fixed training grid. gllvm used its fitted log-likelihood and sjSDM its checked penalised training log-likelihood. The 33 flagged gllvm fits and two flagged sjSDM fits had an acceptable selected estimate but did not meet this agreement check. In each of the 14 failed gllvm combinations, no start met the numerical acceptance rules, which required reported convergence, finite gradients and a maximum absolute gradient below 0.01. Thus, "failed" here does not necessarily mean that the software crashed; it means that none of its attempts supplied an estimate we could accept under the declared rules.

All **121 flagged fits remain in the scored results** and are marked as unresolved in the comparisons below. Failed combinations remain in the counts and archive but cannot contribute a prediction error; paired comparisons require estimates at both settings. Read the number of available communities and their diagnostic status alongside each average, because leaving out difficult failed cases can affect that average. We did not keep refitting until favourable results appeared, or use the known truth to choose a start. sjSDM has fewer attempted combinations because it did not enter the trait experiment. These counts describe the specified datasets, settings and budgets; passing diagnostics does not establish ecological accuracy or a general ranking of the packages.

### What ecological changes did we simulate?

These are the **true generating curves** for the first species in the first simulated community. They show what the ecological changes mean. They are not fitted results. The other environmental predictor is held at zero, and hidden site conditions are averaged over.

```{r extension-generating-curves, fig.height=5}
extension$generating_curves |>
  as_tibble() |>
  filter(scenario %in% c("baseline", "rare", "curved")) |>
  ggplot(aes(environment, probability, colour = scenario)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x = "Environmental gradient 1, in original simulation units",
    y = "True probability of occurrence",
    colour = "Simulation",
    caption = "Known generating truth for species_01, community 1; no model estimates in this figure."
  ) +
  theme_bw()
```

The curved example has a preferred environmental value. Giving a straight-response model more sites cannot make it express that shape. We therefore fit the curved datasets twice: once with the two ordinary predictors, and once with an additional squared-gradient predictor. Each package receives the same set of predictors in each comparison. This separates **too little information** from **an unsuitable response shape**.

```{r extension-curved-design-example}
# Construct the curve before learning centring and scaling from training sites.
training_predictors <- as_tibble(extension$curve_training_raw) |>
  mutate(environment_1_squared = environment_1^2)

# Repeat the same operation at test sites, then use the training transformation.
test_predictors <- as_tibble(extension$curve_test_raw) |>
  mutate(environment_1_squared = environment_1^2)
```

A pair of strongly correlated environmental measurements creates a different difficulty. If warm sites are nearly always dry, their observations provide little information about whether warmth or dryness explains a species' response. Predictions within that same setting can look reasonable even when the individual environmental effects remain uncertain.

```{r extension-correlated-environment, fig.height=4}
extension$correlated_environment |>
  as_tibble() |>
  ggplot(aes(environment_1, environment_2)) +
  geom_point(alpha = 0.45) +
  labs(
    x = "Environmental gradient 1",
    y = "Environmental gradient 2",
    caption = "Actual simulated training conditions; generating correlation = 0.85."
  ) +
  theme_bw()
```

### Compare errors one community at a time

The vertical axis below is the average absolute difference between estimated and true new-site probabilities, in percentage points. Zero means perfect recovery of the probability. Each connected pair represents the **same simulated community**, fitted at the two sample sizes. A downward line means that adding sites reduced error in that community. Differences among the ten lines show how much the answer depends on the dataset.

A fit whose diagnostics remain unresolved is marked separately. An unsuccessful fit has no error estimate and stays in the status table. Neither kind of difficulty should disappear from a package comparison.

```{r extension-sample-size-errors, fig.height=8}
if (nrow(extension$overall) > 0) {
  ecological_errors <- as_tibble(extension$overall) |>
    filter(scenario != "traits") |>
    mutate(
      sites = factor(n_sites, levels = c(100, 300)),
      diagnostic_status = if_else(diagnostic_pass, "Passed checks", "Unresolved"),
      comparison = paste(scenario, response, sep = ": ")
    )

  ggplot(ecological_errors, aes(sites, mae_pp, group = replicate)) +
    geom_line(alpha = 0.4) +
    geom_point(aes(shape = diagnostic_status), size = 2) +
    facet_grid(comparison ~ package) +
    labs(
      x = "Training sites",
      y = "Average absolute error (percentage points)",
      shape = "Fit diagnostics",
      caption = "Each line is one independent community. The table above records missing fits; incomplete runs are not a benchmark."
    ) +
    theme_bw()
}
```

```{r extension-paired-summary}
if (nrow(extension$overall) > 0) {
  paired_errors <- as_tibble(extension$overall) |>
    filter(scenario != "traits") |>
    select(package, scenario, response, replicate, n_sites, mae_pp, diagnostic_pass) |>
    tidyr::pivot_wider(
      names_from = n_sites,
      values_from = c(mae_pp, diagnostic_pass),
      names_expand = TRUE
    )

  if (all(c("mae_pp_100", "mae_pp_300") %in% names(paired_errors))) {
    paired_summary <- paired_errors |>
      filter(!is.na(mae_pp_100), !is.na(mae_pp_300)) |>
      group_by(package, scenario, response) |>
      summarise(
        paired_communities = n(),
        pairs_with_unresolved_diagnostics = sum(!diagnostic_pass_100 | !diagnostic_pass_300),
        error_100_sites = mean(mae_pp_100),
        error_300_sites = mean(mae_pp_300),
        change_in_error = mean(mae_pp_300 - mae_pp_100),
        .groups = "drop"
      )

    knitr::kable(paired_summary, digits = 2)
  }
}
```

A negative change means lower error with 300 sites. These averages use only communities with estimates at both site counts, including any marked unresolved fits. Read the number of paired communities and the diagnostic counts alongside every average. An omitted failed fit is not evidence that its prediction error was zero.

```{r extension-fitted-curves, fig.height=7}
if (nrow(extension$fitted_curves) > 0) {
  extension$fitted_curves |>
    as_tibble() |>
    ggplot(aes(environment)) +
    geom_line(aes(y = truth), colour = "black", linetype = "dashed", linewidth = 0.8) +
    geom_line(aes(y = estimate, colour = response), linewidth = 0.8) +
    facet_grid(n_sites ~ package) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    labs(
      x = "Environmental gradient 1",
      y = "Occurrence probability",
      colour = "Fitted response",
      caption = "Species_01, community 1. Dashed black: generating truth. Coloured lines: fitted marginal probabilities. Rows show training-site counts."
    ) +
    theme_bw()
}
```

Do not treat the thousands of species-by-site predictions as thousands of independent experimental repetitions. The replication unit here is a simulated community. For a formal sample-size summary we need matched completed fits at both site counts and must report which communities, if any, could not be compared.

## 13. Do traits explain species responses, and do they improve prediction?

A species' environmental effect describes how its distribution changes along a gradient. A **trait effect** asks why those environmental responses differ among species. For example, do drought-tolerant species respond less negatively to increasingly dry conditions?

This experiment contains two measured traits. The first truly changes the environmental responses. The second has **no generating effect**. Individual species also differ for reasons that these measured traits do not explain. Knowing the simulation truth lets us distinguish missing a real effect from finding an apparent effect for an irrelevant trait.

There are two separate ways to collect more information:

- **More sites, 100 versus 300:** better observations of each species' environmental response.
- **More species, 10 versus 30:** more species with which to estimate the relationship between traits and environmental responses.

The larger community contains the original ten species. Both site counts are used at both species counts. For every dataset, we fit models **with and without the measured traits**. In gllvm we retain species random slopes in both arms, so the comparison does not also switch on an extra allowance for unexplained species differences.

occJSDM, gllvm and Hmsc have suitable trait-model interfaces. The inspected sjSDM fork does not have an equivalent trait-fitting interface, so its absence from this part is not a failed trait-recovery result.

```{r extension-trait-prediction-errors, fig.height=7}
if (nrow(extension$overall) > 0) {
  trait_errors <- as_tibble(extension$overall) |>
    filter(scenario == "traits") |>
    mutate(
      trait_information = if_else(use_traits, "Traits supplied", "Traits omitted"),
      design = paste(n_sites, "sites x", n_species, "species")
    )

  if (nrow(trait_errors) > 0) {
    ggplot(trait_errors, aes(trait_information, mae_pp, group = replicate)) +
      geom_line(alpha = 0.4) +
      geom_point(aes(shape = diagnostic_pass), size = 2) +
      facet_grid(design ~ package) +
      labs(
        x = NULL,
        y = "Average absolute error (percentage points)",
        shape = "Passed fit checks",
        caption = "Paired fits use identical observations. Lower error means probabilities are closer to their simulated truth."
      ) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 25, hjust = 1))
  }
}
```

Improved prediction and clear evidence of a trait relationship are different outcomes. A model can predict reasonably while remaining uncertain about why species differ. Conversely, a detectable trait relationship need not greatly improve predictions when each species already has plenty of observations.

The next figure subtracts the known true trait effect from each estimate. The dashed black line therefore means **accurate estimation**, not necessarily **no trait effect**. The dotted red line means an estimated effect of zero. For the irrelevant trait, those references coincide because its true effect is zero. An interval crossing the black line includes the truth; an interval crossing the red line includes no effect. These are different questions for a trait whose generating effect is nonzero.

```{r extension-trait-coefficient-recovery, fig.height=9}
if (nrow(extension$traits) > 0) {
  matched_trait_effects <- as_tibble(extension$traits) |>
    filter(link == "logit") |>
    mutate(
      coefficient_error = estimate - truth,
      lower_error = lower - truth,
      upper_error = upper - truth,
      design = paste(n_sites, "sites x", n_species, "species"),
      relationship = paste(trait, environment, sep = " / ")
    )

  no_effect_references <- matched_trait_effects |>
    distinct(relationship, design, truth) |>
    mutate(no_effect_on_error_scale = -truth)

  ggplot(matched_trait_effects, aes(factor(replicate), coefficient_error, colour = package)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_hline(
      data = no_effect_references,
      aes(yintercept = no_effect_on_error_scale),
      colour = "#D55E00", linetype = "dotted"
    ) +
    geom_pointrange(
      aes(ymin = lower_error, ymax = upper_error),
      position = position_dodge(width = 0.5)
    ) +
    facet_grid(relationship ~ design, scales = "free_y") +
    labs(
      x = "Independent simulated community",
      y = "Estimated trait effect minus its true value",
      colour = "Package",
      caption = "Dashed black: estimate equals truth. Dotted red: estimated effect is zero. Coefficient differences are not probability percentage points."
    ) +
    theme_bw()
}
```

These coefficient comparisons undo both environmental and trait standardisation. Hmsc's probit coefficients have different units from the simulation's logit coefficients, so they are not overlaid on the same numerical truth here. Hmsc remains in the probability comparison above, where estimates and truth share the same meaning. gllvm's saved records retain its standard-error warnings; passing the objective and gradient checks does not establish that every uncertainty estimate is reliable. Its coefficient table records the true direction or absence of each relationship and explicitly identifies the different link scale.

Ten independent communities are useful for an exploratory teaching comparison. They do not provide precise estimates of how often an interval misses a true effect or incorrectly excludes zero for an irrelevant trait. We retain the individual results and the diagnostic failures instead of turning a small experiment into a claim that one package is generally best.

## 14. Across communities, are estimates biased?

The September fits can answer more questions without being fitted again. We now use the same saved results to distinguish **bias**, **typical error** and **interval coverage**. All four packages remain in this comparison. These calculations extend the ten-community experiment; they do not add new simulated communities.

An estimate's **signed error** is its estimate minus truth. Averaging signed errors over independent simulated communities estimates bias: persistent overestimation is positive, and persistent underestimation is negative. Errors in opposite directions can cancel, so we also report **root mean squared error (RMSE)**. RMSE squares each error before averaging and then takes a square root. It describes error size and gives larger errors more weight than mean absolute error does.

```{r calibration-data}
calibration <- readRDS("teaching-data/lesson-4-calibration.rds")
stopifnot(calibration$complete)

prediction_calibration <- as_tibble(calibration$prediction) |>
  mutate(package = factor(package, levels = package_order))

prediction_summary <- prediction_calibration |>
  group_by(package, scenario, n_sites, n_species, response, use_traits) |>
  summarise(
    communities = n(),
    flagged = sum(!diagnostic_pass),
    bias_pp = 100 * mean(bias),
    bias_mcse_pp = if (n() > 1) 100 * sd(bias) / sqrt(n()) else NA_real_,
    rmse_pp = 100 * sqrt(mean(rmse^2)),
    .groups = "drop"
  )

prediction_summary |>
  filter(scenario == "baseline") |>
  select(package, n_sites, communities, flagged, bias_pp, bias_mcse_pp, rmse_pp) |>
  knitr::kable(digits = 2, col.names = c(
    "Package", "Sites", "Communities", "Flagged",
    "Bias (points)", "Bias MCSE (points)", "RMSE (points)"
  ))
```

Here the target is each species' occurrence probability at the 300 independent test sites, averaged over unknown hidden conditions. Each community receives equal weight. The **Monte Carlo standard error (MCSE)** describes how much the estimated average bias varies because we simulated only ten communities. It is not a species' posterior standard error. The thousands of species-by-site cells do not turn this into thousands of independent repetitions. RMSE is the square root of the average community mean squared error, not the average of community RMSEs.

```{r calibration-error-comparison, fig.height=8}
prediction_calibration |>
  filter(scenario != "traits") |>
  mutate(
    condition = paste(scenario, response, sep = ": "),
    diagnostic = if_else(diagnostic_pass, "Passed checks", "Unresolved"),
    sites = factor(n_sites)
  ) |>
  select(package, replicate, condition, diagnostic, sites, bias, rmse) |>
  pivot_longer(c(bias, rmse), names_to = "measure", values_to = "error") |>
  mutate(measure = recode(measure, bias = "Signed error", rmse = "RMSE")) |>
  ggplot(aes(package, 100 * error, colour = sites, shape = diagnostic)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_point(position = position_jitterdodge(jitter.width = .15, dodge.width = .6, seed = 25), alpha = .75) +
  facet_grid(condition ~ measure, scales = "free_y") +
  labs(x = NULL, y = "Error (percentage points)", colour = "Training sites", shape = "Diagnostics",
    caption = "Each point is one community. Flagged fits remain visible; failed fits have no score.") +
  theme_bw() + theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

Small signed errors alongside appreciable RMSE mean that errors cancel, not that individual estimates are precise. Compare the distributions of community results, rather than choosing a winner from a small difference between averages. The failure counts in section 12 still apply. Where a package failed on some communities, its average describes only the communities with a scored fit.

## 15. Do 95% intervals contain the truth?

**Coverage** asks how often an interval includes the generating truth across repeated datasets. For example, an interval from 0.2 to 0.5 covers a true occurrence probability of 0.4 but misses a truth of 0.7. A narrow interval can be confidently wrong; a very wide interval can cover the truth while saying little. We therefore show **coverage and interval width together**.

Bayesian credible intervals and approximate frequentist confidence intervals have different definitions. Repeated simulation lets us measure the coverage of either procedure. Bayesian intervals are not guaranteed to have exactly 95% coverage under the ecological truth distribution used here.

### Which saved results support this calculation?

- **All four packages:** point errors for test-site probabilities and five shared environmental settings.
- **occJSDM and Hmsc:** 95% equal-tailed intervals for marginal occurrence probabilities, calculated from every saved global parameter draw.
- **occJSDM:** intervals for environmental coefficients, where the fitted response contains the generating terms and the logit coefficient definitions match.
- **occJSDM and gllvm:** existing intervals for the measured trait effects on their matching logit scale.
- **gllvm and sjSDM probability intervals, and their ordinary environmental-coefficient intervals:** unavailable from the records saved for this experiment. Missing intervals are recorded as unavailable, not as zero coverage.

Hmsc fitted a probit model to logit-generated data. Its probabilities can be compared with the true probabilities, but its raw coefficients do not have the same numerical truth as the logit coefficients. For coefficient recovery we restrict the calculation to conditions where the fitted response includes the generating terms for every species. The straight-response fit to the curved scenario is therefore omitted from the coefficient tables as a whole, including the species whose generating responses remain straight. Its probability errors and coverage still include every species. This study measures performance under the existing logit-generating design. A probit-generating arm is still needed for a balanced comparison of that modelling choice.

### Probability intervals at five shared environments

To keep the interval calculation inspectable, we evaluate five fixed settings: both environmental gradients at zero, then each gradient at -1 and +1 while the other remains zero. These are original simulation units, before training-data standardisation. Every package and community uses the same settings. Hidden site conditions are integrated out **within each posterior draw**, and the interval is then calculated across those marginal-probability draws. These are intervals for the average probability given the measured environment, not intervals for a particular site's hidden conditions or for a future binary observation.

```{r calibration-grid}
knitr::kable(calibration$grid, row.names = TRUE)
```

The probability coverage calculation therefore uses a different evaluation set from the 300-site point-error table above. Its mean, lower endpoint and upper endpoint were checked for numerical integration accuracy. It uses every archived draw, including all four chains. The original fitting diagnostics remain attached; post-processing cannot repair an unconverged chain.

```{r calibration-probability-summary}
probability_summary <- as_tibble(calibration$summary) |>
  filter(target == "Marginal probability", scenario == "baseline", population == "All scored fits") |>
  mutate(package = factor(package, levels = package_order)) |>
  arrange(package, n_sites)

probability_summary |>
  transmute(
    Package = package, Sites = n_sites, Communities = communities,
    Flagged = flagged_communities, `Communities with intervals` = interval_communities,
    `Coverage (%)` = 100 * coverage, `Coverage MCSE (points)` = 100 * coverage_mcse,
    `Mean width (points)` = 100 * width
  ) |>
  knitr::kable(digits = 1)
```

`NA` means that this archive cannot supply the interval comparison. Within each community we average coverage and width over its species and the five settings; we then average those community summaries equally. Coverage MCSE is calculated from the variation among the community summaries. It is not a binomial standard error that treats every species-setting pair as independent. With ten communities, the result is exploratory and the MCSE itself is uncertain.

```{r calibration-probability-intervals, fig.height=8}
as_tibble(calibration$per_community) |>
  filter(target == "Marginal probability", scenario != "traits", intervals > 0) |>
  mutate(
    condition = paste(scenario, response, sep = ": "),
    diagnostic = if_else(diagnostic_pass, "Passed checks", "Unresolved"),
    sites = factor(n_sites)
  ) |>
  select(package, replicate, condition, diagnostic, sites, coverage, width) |>
  pivot_longer(c(coverage, width), names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure, coverage = "Coverage (%)", width = "Mean width (points)")) |>
  ggplot(aes(package, 100 * value, colour = sites, shape = diagnostic)) +
  geom_hline(data = data.frame(measure = "Coverage (%)"), aes(yintercept = 95), linetype = 2, inherit.aes = FALSE) +
  geom_point(position = position_jitterdodge(jitter.width = .15, dodge.width = .6, seed = 25), alpha = .8) +
  facet_grid(condition ~ measure, scales = "free_y") +
  labs(x = NULL, y = NULL, colour = "Training sites", shape = "Diagnostics",
    caption = "Each point summarises one community. The dashed line marks nominal 95% coverage.") +
  theme_bw()
```

The straight-response fits to curved truth illustrate why width matters. At 300 sites, their measured probability coverage is `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "occJSDM" & scenario == "curved" & response == "linear" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` for occJSDM and `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "Hmsc" & scenario == "curved" & response == "linear" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` for Hmsc. Including the quadratic term gives `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "occJSDM" & scenario == "curved" & response == "quadratic" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)` and `r sprintf("%.1f%%", 100 * subset(calibration$summary, package == "Hmsc" & scenario == "curved" & response == "quadratic" & n_sites == 300 & target == "Marginal probability" & population == "All scored fits")$coverage)`, respectively. More data can narrow intervals around an unsuitable response shape; it does not supply the missing curve. These numbers describe the five settings and ten communities tested, including flagged fits.

The baseline is also not a substitute for checking difficult conditions. The rare-species and correlated-environment panels show lower occJSDM coverage than its baseline. This records a pattern to investigate with further replication and diagnostic work; it does not by itself identify a software defect or establish an overall package ranking.

All scored fits are retained above. A companion summary restricted to fits that passed the original diagnostics is also saved. Compare it with the full summary as a sensitivity check; excluding difficult fits can select easier communities, so the restricted summary is not a corrected package ranking.

```{r calibration-diagnostic-sensitivity}
as_tibble(calibration$summary) |>
  filter(target == "Marginal probability", scenario == "baseline", intervals > 0) |>
  transmute(Package = package, Sites = n_sites, Fits = population,
    Communities = interval_communities, `Coverage (%)` = 100 * coverage,
    `Mean width (points)` = 100 * width) |>
  knitr::kable(digits = 1)
```

### Environmental coefficients and trait relationships

For compatible coefficient targets we undo predictor scaling and centring. For the intercept, the adjustment is made jointly with each draw's slopes before taking quantiles. Rescaling a marginal intercept interval alone would discard the relevant dependence.

```{r calibration-environmental-effects}
as_tibble(calibration$summary) |>
  filter(target == "Environmental coefficient", scenario == "baseline", n_sites == 100,
    population == "All scored fits") |>
  transmute(Package = package, Coefficient = term, Communities = communities,
    Bias = bias, RMSE = rmse, `Coverage (%)` = 100 * coverage, `Mean width` = width) |>
  knitr::kable(digits = 3)
```

The coefficient table reports the same conditional ecological coefficient target as the simulator; it does not compare the orientation of latent factor axes. The probability comparison remains the shared target across all four packages. Hmsc is absent from this coefficient table because its link differs, not because it failed to supply estimates.

```{r calibration-trait-effects}
as_tibble(calibration$summary) |>
  filter(target == "Trait coefficient", n_sites == 100, n_species == 10,
    population == "All scored fits") |>
  transmute(Package = package, Relationship = term, Communities = communities,
    Flagged = flagged_communities, Bias = bias, `Coverage (%)` = 100 * coverage,
    `Mean width` = width) |>
  knitr::kable(digits = 3)
```

This displayed trait slice uses 100 sites and ten species; the saved summary also contains the 300-site and thirty-species cases. For the irrelevant trait, truth is zero, so an interval that misses truth also excludes zero incorrectly. For a genuinely nonzero effect, covering truth and excluding zero answer different questions. gllvm's original standard-error warnings and unavailable fits still matter; these are approximate intervals from the selected fits, not a guarantee of calibration.

The compact bundle contains per-community summaries and individual grid/coefficient records, so these tables can be filtered without accessing the large fitting archive. For example:

```{r calibration-inspect}
as_tibble(calibration$per_community) |>
  filter(package == "occJSDM", target == "Marginal probability",
    scenario == "baseline", n_sites == 100) |>
  select(replicate, diagnostic_pass, bias, rmse, coverage, width)
```

No model was refitted for these sections. A larger coverage study should add independent communities, complete the gllvm/sjSDM uncertainty methods, and include a probit-generating arm under a declared fitting and diagnostic protocol. Preserve these ten communities and their failures as the original experiment. The July/August occJSDM-only study predates model fixes and remains historical evidence; it is not pooled with these September results.
