---
title: "occJSDM: quickstart and lesson guide"
output:
  rmarkdown::html_vignette:
    toc: true
    css: teaching.css
vignette: >
  %\VignetteIndexEntry{occJSDM: quickstart and lesson guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: none
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, comment = "#>")

lesson_link <- function(name) {
  is_html <- knitr::pandoc_to() %in% c("html", "html4", "html5")
  paste0(name, if (is_html) ".html" else ".md")
}
```

occJSDM estimates species occurrence while allowing for imperfect field collection, PCR detection and false positives. It can also fit a JSDM to directly observed presence/absence data. The teaching lessons show the R code and compare the results with the truth used to simulate their data.

## Choose a lesson

- [Lesson 0: Create and explore a simulated survey](`r lesson_link("occJSDM-lesson-0")`). Optional preparation: sites, field samples, primers, PCR replicates, environmental covariates, traits and maps of the known truth.
- [Lesson 1: Fit the model and compare its answers with truth](`r lesson_link("occJSDM-first-lesson")`). The main starting point: fit the model, understand occupancy error, and inspect actual examples of weak true detections, strong true detections and false positives.
- [Lesson 2: Spatial landscapes and dispersal](`r lesson_link("occJSDM-lesson-2")`). A planned lesson on smooth environmental gradients, residual spatial structure and a separate dispersal simulation. It is not yet a worked spatial validation.
- [Lesson 3: Understand the model's outputs by comparing them with truth](`r lesson_link("occJSDM-lesson-3")`). Environmental responses, trait effects, species associations, ordination, variation partitioning and detection effort. It includes a function index and explains why a real simulated effect can remain uncertain.

You can read Lesson 3 after Lesson 1; the spatial lesson is not a prerequisite. The earlier output tour has been replaced by Lesson 3. The old `sampledata` and `sampleresults` objects remain available, but their generating truth is not stored with them, so the lessons use a complete matching simulation and fit instead.

## A minimal fitting example

From the repository's `vignettes` directory, the following optional code fits the same non-spatial survey used in the lessons. It is shown without running during knitting because MCMC takes time. Lessons 1 and 3 render saved, checked results and show their matching truth.

```{r quickstart, eval=FALSE}
library(occJSDM)

lesson <- readRDS("teaching-data/nonspatial-lesson.rds")
survey_data <- lesson$input$sim$data_list
known_truth <- lesson$input$sim$true_params

set.seed(20260921)

fitmodel <- runOccJSDM(
  data = survey_data,
  listParams = list(n_factors = 2, n_lattrait = 1),
  threshold = 1,
  occCovariates = c("X_psi.EnvCov.1", "X_psi.EnvCov.2"),
  collCovariates = "X_theta",
  spatCovariates = NULL,
  MCMCparams = list(nchain = 4, nburn = 3000, niter = 6000, nthin = 1),
  summarisedLatentPresences = TRUE
)
```

`survey_data` is what the model receives. `known_truth` is kept separately for evaluation. `n_factors` controls the hidden site factors, whereas `n_lattrait` controls the unmeasured species-trait structure. These are different parts of the model.

For reproduction with the original source and complete saved fits, follow `dev/simstudy/vignette-lesson/README.md`. Using the same seed with a different package revision does not guarantee the original results. The lessons record source and fit hashes and keep rendering separate from fitting.

## What to inspect next

Start with the diagnostics and probability comparisons in Lesson 1, then the coefficient, trait and community-output figures in Lesson 3. An interval crossing zero does not demonstrate that an ecological effect is absent; an interval excluding zero does not demonstrate that its magnitude is accurate.

For a particular function, use the index at the end of Lesson 3 or the package's R help. Genuine new-site prediction, spatial validation and comparisons with other JSDM packages require their own matched examples and are not established by the current non-spatial lessons.
