---
title: "occJSDM"
output: 
  rmarkdown::html_vignette:
    fig.width: 7
    fig.height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{occJSDM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  comment = "#>"
)

options(warning = -1)
```

```{r setup, include = FALSE}
# library(occPlus)
devtools::load_all() # change to library(occPlus) when package finished
```

The model can be run using the `runOccPlus` function.

The arguments of the function are:

- data: which we explain below;
- listParams: a list of the number of ordination factors used for
  computing occupancy probabilities and of the number of unobserved
  traits;
- occCovariates: vectors of covariates predicting occupancy
  probabilities;
- collCovariates: vectors of covariates predicting eDNA collection (=
  detection) probabilities;
- spatCovariates: vectors of covariates (e.g. X, Y) used to construct a
  spatial random field capturing residual spatial structure in occupancy
  that is not explained by occCovariates;
- traitsMatrix: a matrix of species traits (rows = species, columns =
  traits) used to explain how species' occupancy responses to
  occCovariates vary as a function of their traits;
- MCMCparams: a list specifying the MCMC settings, including nchain
  (number of chains), nburn (number of burn-in iterations to discard),
  niter (number of post-burn-in iterations to keep), and nthin (thinning
  interval, i.e. how many iterations to skip between saved samples).

## Data preparation

The data object is a list of two elements: `info` and `OTU`.

```{r}
data <- data
head(data$info)
```

`data$info` is a data.frame with $N$ rows, where $N$ is the total number
of samples analyzed and columns:

- `Site`: name (or index) of the site in which this replicate was
  collected;
- `Sample`: name (or index) of the sample from which this replicate was
  analyzed;
- `Primer`: name (or index) of the primer used in this replicate;
- `X_psi.n` are the occupancy covariates (e.g. elevation);
- `X_theta.n` are the eDNA collection covariates (e.g. water volume per
  sample).

In this dataset, there are $100$ sites and, in the balanced case, $m=3$
sample replicates per site and $L=3$ primers. [From this data
structure]{.underline}, [occJSDM can infer that there are $K=2$ PCR
replicates per sample, so a fully balanced dataset would have had
$3 \times 3 \times 2 = 1800$ rows (i.e. observations). However, $3$ of
the $300$ samples (one each at three different sites) are dropped to
simulate missing observations, leaving $297$ samples and $1{,}782$ total
replicates. This illustrates that occJSDM can work with unbalanced study
designs, such as sites with a missing samples.]{.underline}

We can see this imbalance directly by counting the number of rows per
site: most sites have $18$ rows ($3$ samples $\times$ $3$ primers
$\times$ $2$ PCR replicates), but the three affected sites have only
$12$.

```{r}
data$info |> count(Site)
```

```{r}
head(data$OTU)
```

`data$OTU` is a matrix of dimension $N \times S$, where $N$ is again the
number of samples (in the same order as `data$info`), and $S$ is the
number of species. Each element in the matrix is the number of reads of
the species in that replicate, [which will be converted to
detection/non-detection (i.e. 1/0)]{.underline}.

## Model fitting

Next, we can run the model by typing (leaving out `spatCovariates` and
`traitsMatrix` for now)

```{r, eval = FALSE}
fitmodel  <- runOccPlus(data,
                        listParams = list(n_factors = 2, 
                                          gt = 0),
                        occCovariates = c("X_psi.1", "X_psi.2"),
                        collCovariates = c("X_theta"),
                        MCMCparams = list(nchain = 2,
                                          nburn = 500,
                                          niter = 500)
                        );beepr::beep()
```

The names of the covariates in `occCovariates` and `collCovariates`
should match the column names in `data$info`.

## Output

There are several outputs we can obtain:

### Covariate coefficients

First, we provide two functions for returning collection and occupancy
covariate coefficients. The output is returned in the form (*niter* x
*numberOfCovariates* x *numberOfSpecies*).

```{r}
collCovariatesOutput <- returnCollectionCovariates(fitmodel)
str(collCovariatesOutput)
```

```{r}
occCovariatesOutput <- returnOccupancyCovariates(fitmodel)
str(occCovariatesOutput)
```

We provide two functions for plotting collection and occupancy covariate
coefficients:

```{r}
plotCollectionCovariates(fitmodel, idx_species = 1:10, 
                         covName = "X_theta")
```

```{r}
plotOccupancyCovariates(fitmodel, idx_species = 1:10, 
                        covName = "X_psi.1")

plotOccupancyCovariates(fitmodel, idx_species = 1:10, 
                        covName = "X_psi.2")
```

### Collection, Detection, and False-Positive rates

Second, we can plot collection, detection, and false-positive rates for
Stage 1 (eDNA collection in the field) and Stage 2 (eDNA detection in
the lab).

```{r Stage 1}
plotStage1FPRates(fitmodel, idx_species = 1:10) + ylim(c(0,1)) # ideally, these are low
plotCollectionRates(fitmodel, idx_species = 1:10) # ideally, these are high
```

```{r Stage 2}
plotStage2FPRates(fitmodel, idx_species = 1:10) + ylim(c(0,1)) # ideally, these are low
plotDetectionRates(fitmodel, idx_species = 1:10) # ideally, these are high. 
```

Note that the Stage 2 detection rates are calculated separately for each
primer and species, which allows one to determine how well each primer
amplifies each species.

And false positives rates at the first stage (field stage):

```{r}
plotStage1FPRates(fitmodel, idx_species = 1:20)
```

And at the second stage (lab stage):

```{r}
plotStage2FPRates(fitmodel, idx_species = 1:20)
```

### Occupancy rates

To compute occupancy probabilities at each site and for each species, we
use the following function.

```{r, eval = T}
occProbs <- computeOccupancyProbs(fitmodel)
str(occProbs)
```

The output is stored in the form (*iterations* x *sites* x *species*).

### Read intensity

`occPlus` estimates a distribution of reads for true and false
positives. We can plot the two distributions via:

```{r, fig.width=7, fig.height=4.5, warnings = FALSE}
plotReadIntensity(fitmodel)
```

### Covariance matrix

We can plot the posterior median of the residual correlation matrix as

```{r}
plotCorrelationMatrix(fitmodel)
```

We can show only the positively and negatively significant elements
(positive in red and negative in blue)

```{r}
plotSigElementsCorMatrix(fitmodel)
```

We note that this is the \textit{residual} covariance matrix,
\textbf{after} accounting for occupancy covariates but accounting for
ordination covariates.
