---
title: "Lesson 0 (optional): Create and explore a simulated survey"
output:
  rmarkdown::html_vignette:
    toc: true
    css: teaching.css
    fig_width: 8
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Lesson 0 (optional): Create and explore a simulated survey}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r document-settings, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, collapse = FALSE, comment = "#>",
  fig.width = 8, fig.height = 5, dpi = 150, dev = "png"
)
source("lesson-links.R")

```

## Where this lesson fits

We will learn occJSDM using a community whose true distribution is known. A real survey never reveals which sites a species truly occupies, so it cannot show whether a model's answers are right. A simulation keeps those answers, so every estimate the model makes can be checked against the truth that generated the data. occJSDM estimates which species occupy which sites from eDNA surveys. It allows for false negatives (from imperfect field collection, imperfect PCR detection or both) and for false positives (from contamination in the field, in the lab or both); the [quickstart](occJSDM.md) shows a first fit on the package's example data. This lesson creates and explores the survey, and [Lesson 1](occJSDM-lesson-1.md) fits the model.

You can skip Lesson 0 and start Lesson 1 with the supplied dataset. Read this lesson if you want to understand where that dataset came from, change the simulation settings, or learn how to prepare the tables and figures yourself.

This survey has no spatial structure. Its sites have coordinates, which let us draw maps, but location does not generate its environmental values or enter its fitted model. That makes it a good place to start: the model's answers can be checked without also asking whether it has captured a spatial pattern. The spatial lesson deliberately changes that. Mapping data and fitting a spatial model are different things.

## Load the teaching data and the R tools

The teaching data are in the occJSDM repository, not in the installed package, so first clone or download [the repository from GitHub](https://github.com/AlexDiana/occJSDM). Run the code with the repository's `vignettes` directory as your working directory. Knitting this document uses that directory automatically. In RStudio, opening the `.Rmd` file and choosing **Session \> Set Working Directory \> To Source File Location** gives the same starting point for running chunks interactively.

**What this lesson assumes you know.** The code uses base R and the tidyverse: the pipe `|>`, and from dplyr and tidyr the verbs listed below. If any are new, the two chapters of R for Data Science on [data transformation](https://r4ds.hadley.nz/data-transform) and [data tidying](https://r4ds.hadley.nz/data-tidy) teach everything used here in an afternoon. Operations that are unusual, such as joining two tables on species identity, are explained where they appear.

- `select()` and `pull()` to choose columns, or to take one column out as a vector.
- `filter()`, `distinct()`, `arrange()`, `slice_head()` and `slice_sample()` to keep, deduplicate, order and pick rows, including a random pick.
- `mutate()`, `transmute()` and `case_when()` to add or recode columns (`transmute()` keeps only the new ones), and `group_by()`, `summarise()` and `count()` to summarise or count by group.
- `bind_cols()` to put two tables side by side, and `left_join()` and `anti_join()` to combine tables on shared identifiers, or to drop the rows that match.
- `pivot_longer()` to move from one column per measurement to one row per measurement.

```{r load-data, message=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

lesson <- readRDS("teaching-data/nonspatial-lesson.rds")

simulation <- lesson$input$sim
survey_data <- simulation$data_list
known_truth <- simulation$true_params

format_percent <- function(probability, digits = 1) {
  paste0(formatC(100 * probability, format = "f", digits = digits), "%")
}
```

`format_percent()` prints probabilities as percentages in the text. Here is what the saved file holds.

```{r lesson-contents}
names(lesson)
str(survey_data, max.level = 1)
names(known_truth)
```

`lesson` is the saved file that Lesson 1 also reads. `input` holds the simulation, and `cases` the four detection examples that Lesson 1 selected; most of the other components summarise Lesson 1's model fits, which this lesson does not use. `survey_data` is what the two-stage model receives: the three tables `info`, `OTU` and `traits`, which "Understand the three input objects" describes. `known_truth` is what the simulation knew and the model never sees. `z_true` records whether each species truly occupies each site, and `w_true` whether its DNA truly entered each field sample. `jsdmParams_true` holds the ecological model's coefficients and each site's true occupancy score for each species, `beta_theta_true` the collection coefficients, and `p_true` and `q_true` the laboratory rates. "Decide how collection and PCR can fail" explains the two-stage process that these truths describe.

occJSDM takes the read counts as a matrix with one column per species, which is how `survey_data$OTU` stores them. To explore the observations we reshape a copy into a long table, with one row per species and PCR.

## Specify the survey, one decision at a time

First give the sampling decisions meaningful names. These are the actual settings of the saved teaching dataset.

```{r survey-design}
n_sites <- 100
n_species <- 10
samples_per_site <- 3
n_primers <- 2
pcrs_per_primer <- 6

n_field_samples <- n_sites * samples_per_site
n_pcr_observations <- n_field_samples * n_primers * pcrs_per_primer

tibble(
  quantity = c("Sites", "Species", "Field samples", "PCR reactions"),
  number = c(n_sites, n_species, n_field_samples, n_pcr_observations)
)
```

There are `r n_field_samples` field samples and `r format(n_pcr_observations, big.mark = ",")` PCR reactions. As in metabarcoding, each reaction yields a read count for every species, so there are not separate PCRs for each species. Six PCR replicates means **six per primer per field sample**, not six divided between the two primers.

These numbers are a choice made for teaching. One hundred sites and ten species make a small community whose every site and species the lessons can display. Three field samples per site matches the quickstart's `sampledata`. Field replication is what lets the model weigh a sample's positive against the site's other samples, and [Lesson 1's appendix](occJSDM-lesson-1.md#why-the-survey-has-three-field-samples-per-site) records why two samples per site were not enough. Two primers with six PCRs each give every sample `r n_primers * pcrs_per_primer` PCRs, many repeated laboratory results from which to learn the laboratory rates. On your own survey these numbers are whatever your field and laboratory design produced; you do not choose them for occJSDM.

The simulator uses shorter argument names. This list translates our decisions into its interface. `M` and `K` are vectors rather than single numbers. `M` holds the number of field samples at each site, and `K` the number of PCRs for each combination of field sample and primer, running through the samples in order with each sample's primers together. Vectors are what allow unequal replication, which "Unequal numbers of field samples" prepares; here every value of `K` is six, so their order does not matter. A **trait** is a measured property of each species, such as body size. An **occupancy covariate** describes the habitat at a site, such as water temperature. A **collection covariate** describes how each field sample was taken, such as the volume of water filtered: it changes how likely a sample is to capture DNA, not whether the species is there.

```{r simulator-settings}
survey_settings <- list(
  n = n_sites,
  S = n_species,
  g = 2,                         # Two measured traits per species.
  M = rep(samples_per_site, n_sites),
  P = n_primers,
  K = rep(pcrs_per_primer, n_field_samples * n_primers),
  ncov_psi = 2,                   # Two environmental covariates per site.
  ncov_theta = 1                  # One collection covariate per field sample.
)

str(survey_settings)
```

`M` has `r length(survey_settings$M)` values, one per site, and `K` has `r length(survey_settings$K)`, one per field sample and primer; the other settings are single numbers.

## Decide how collection and PCR can fail

An eDNA survey is a **two-stage process**: DNA is collected in the field, then detected by PCR in the laboratory. Each stage can go wrong in two ways. In the field, a species either occupies a site or not, and each field sample either captures its DNA or not. Collection can miss a species that is present, which is a **false negative**. It can also pick up the species' DNA when the species is absent from the site, for example DNA carried in on equipment from another site, which is a **field-stage false positive**. In the laboratory, a PCR can miss DNA that is in the sample, another false negative. It can also report DNA that is not there, through contamination in the laboratory, which is a **laboratory false positive**. The simulator gives each of these events a probability:

- `theta`, the collection probability: the probability that a field sample captures the species' DNA when the species occupies the site. A collection failure, the field false negative, has probability `1 - theta`.
- `theta0`: the probability that the species' DNA enters a field sample although the species does not occupy the site, the field-stage false positive.
- `p`: the probability that a PCR detects the species when its DNA is in the sample. A PCR failure, the laboratory false negative, has probability `1 - p`.
- `q`: the probability that a PCR reports the species when its DNA is not in the sample, the laboratory false positive.

[The intuition lesson](occJSDM-lesson-intuition.md#what-an-occupancy-model-does) shows the one-stage version of this process, with a single detection probability and no false positives, and [Lesson 1](occJSDM-lesson-1.md#three-questions-three-different-truths) explains the full model.

We deliberately let species differ in how reliably they are collected and detected. For example, primer 1's true-detection event probabilities increase from `r lesson$input$params$p[1, 1]` to `r lesson$input$params$p[1, n_species]` across the ten species. The second primer has somewhat higher probabilities. These are assigned simulation settings, not findings about real primers.

```{r observation-settings}
primer_1_detection <- seq(0.35, 0.85, length.out = n_species)
primer_2_detection <- pmin(0.95, primer_1_detection + 0.10)

observation_settings <- list(
  p = rbind(primer_1_detection, primer_2_detection),
  q = rbind(
    seq(0.025, 0.065, length.out = n_species),
    seq(0.065, 0.025, length.out = n_species)
  ),
  theta0 = seq(0.02, 0.08, length.out = n_species),
  theta_baseline = seq(0.35, 0.75, length.out = n_species),
  mu1 = 5,
  sigma1 = 1,
  mu0 = 1.5,
  sigma0 = 1
)

str(observation_settings)
```

`p` and `q` are matrices with one row per primer and one column per species. `theta0` and `theta_baseline` have one value per species, and the four read-count settings are single numbers. The settings have these meanings:

- `p`: the probability of a laboratory detection event when the species' DNA is in the sample, for each primer and species.
- `q`: the probability of a laboratory false-positive event when the DNA is absent from the sample, for each primer and species.
- `theta0`: the probability that DNA enters a sample although the species is absent from the site, for each species.
- `theta_baseline`: the collection probability `theta` at a collection-covariate value of zero, given that the species occupies the site. The actual collection probability also varies with that covariate. The simulator draws each species' slope on the covariate itself, and `known_truth$beta_theta_true` holds the slopes it drew.
- `mu1` and `sigma1`: the mean and standard deviation, on the log-read scale, of the read count from a true detection event.
- `mu0` and `sigma0`: the same for a laboratory false-positive event.

These settings make OTU_1 the hardest species to detect. It has the lowest collection probability at baseline (`r observation_settings$theta_baseline[1]`) and the lowest PCR detection probability on both primers (`r round(observation_settings$p[1, 1], 2)` and `r round(observation_settings$p[2, 1], 2)`). OTU_10 is the easiest, at `r observation_settings$theta_baseline[n_species]`, `r round(observation_settings$p[1, n_species], 2)` and `r round(observation_settings$p[2, n_species], 2)`. The maps below compare the two species. `pmin(0.95, ...)` caps primer 2's probabilities at 0.95. No value exceeds the cap with these settings, but the cap keeps every probability valid if you raise the first line. The two primers' `q` run in opposite directions across the species, so no species has a high laboratory false-positive rate on both primers. `theta0` rises across the species, so field contamination is commonest for OTU_10.

The simulator turns each event into a read count. It draws a value on the log-read scale, converts it with `round(exp(log_read) - 1)` and sets negative counts to zero. A true detection event therefore gives about `r round(exp(observation_settings$mu1) - 1)` reads at the median, and a laboratory false-positive event about `r round(exp(observation_settings$mu0) - 1)`. A field-stage false positive gives as many reads as a true detection, because its sample really holds the species' DNA. An event can also produce zero reads, when the draw falls below `log(1.5)`. With these settings that almost never happens to a true detection event (`r format_percent(pnorm(log(1.5), observation_settings$mu1, observation_settings$sigma1), 4)` of them), but it happens to `r format_percent(pnorm(log(1.5), observation_settings$mu0, observation_settings$sigma0))` of laboratory false-positive events, about one in `r round(1 / pnorm(log(1.5), observation_settings$mu0, observation_settings$sigma0))`. The fitted model uses positive/negative counts at a threshold, so its `p` and `q` must be compared with the probability of a **threshold-positive result**, not simply with these event probabilities. Here the probability of a threshold-positive laboratory false positive is about `r format_percent(1 - pnorm(log(1.5), observation_settings$mu0, observation_settings$sigma0), 0)` of the `q` set above. Lesson 1 shows the conversion and the resulting truth comparisons.

occJSDM assumes that the survey was run with good field and laboratory practice, so that contamination, and therefore false positives, are infrequent. It expresses this through its **priors**, the starting beliefs about plausible rates that the observations then update. The default priors on `q` and `theta0` are both Beta(1, 20), which has mean 1/(1 + 20), about `r format_percent(1 / (1 + 20))`, and puts 95% of its weight below `r format_percent(qbeta(0.95, 1, 20))`. This simulation's contamination rates are chosen inside that assumption: `q` runs from `r format_percent(min(lesson$input$params$q))` to `r format_percent(max(lesson$input$params$q))`, and `theta0` from `r format_percent(min(lesson$input$params$theta0))` to `r format_percent(max(lesson$input$params$theta0))`. Lesson 1 explains [how good practice enters the model](occJSDM-lesson-1.md#how-does-good-practice-enter-the-model) and [stress-tests the assumption](occJSDM-lesson-1.md#what-changes-if-we-are-less-confident-about-low-contamination) by refitting with priors that expect more contamination.

## Specify the ecological model

The ecological model sets each species' occupancy probability at each site. On the logit scale it adds three parts: the species' baseline, its responses to the site's environmental covariates, and the effects of a few hidden site factors. The inverse logit then turns the sum into a probability. Each species' responses to the environment are built from its traits, measured and unmeasured, plus a remainder of its own. A hidden community factor is a simulated site characteristic that affects several species but is not supplied as a measured environmental column. It need not represent a particular biological mechanism. Because several species respond to the same hidden factors, they occur together more or less often than the measured environment predicts. These are the associations between species that a joint species distribution model estimates, and the number of factors is what the quickstart's `n_factors` sets. [The intuition lesson](occJSDM-lesson-intuition.md#what-a-joint-model-does) explains them by hand.

```{r ecological-settings}
ecological_settings <- list(
  gt = 1,                        # One unmeasured trait per species.
  d = 2,                         # Two hidden community factors.
  ds = 0,                        # Spatial setting, unused here.
  sigma_b = 0.5,                  # Spread of environmental responses beyond traits.
  sigma_bs = 0.5,                 # Spatial setting, unused here.
  sigma_ts = 0.5,                 # Spatial setting, unused here.
  sigma_h = 1,                    # Spread of the hidden site factors.
  sigma_s = 0.5,                  # Spatial setting, unused here.
  l_s = 0.3,                      # Spatial setting, unused here.
  tau = rep(1, n_species),         # Required setting, unused for binary occupancy.
  useSpatField = FALSE
)

str(ecological_settings)
```

`tau` is the only setting with one value per species; every other setting is a single number or a switch. `gt = 1` gives each species one unmeasured trait, a property not in the trait table that also shapes how it responds to the environment. `sigma_b` sets how differently species respond to the same environment beyond what their traits predict: a larger value makes their responses more varied. `sigma_h` sets how much the hidden factors vary between sites: a larger value makes them matter more for occupancy, and so strengthens the associations between species.

The most important choice for this lesson is `useSpatField = FALSE`. There is no simulated spatial field and no dispersal process. We include the unused spatial settings because the simulator accepts this common parameter list; those values do not create a spatial effect while the switch is off.

The simulator draws the environmental values, traits, coefficients and hidden factors using these settings. It then calculates occupancy probabilities and draws actual presence or absence from those probabilities. We do not choose the seed to obtain an especially successful fit: a seed picked because the fit looked good would mislead you about how well the model works.

## Recreate the dataset if you want to

The following optional chunk shows the actual public simulator call. It is **displayed but not run when knitting**, because the figures use the saved simulation. To recreate the data yourself, run the chunk with occJSDM installed. It does not fit a model. `model = "two_stage"` asks for PCR read counts. The simulator can also produce `"occupancy"` data, with one record per field sample and no PCR stage, and `"binary"` presence/absence data, like the presence/absence case described in the quickstart.

```{r simulate-survey, eval=FALSE}
library(occJSDM)

set.seed(20260919)

new_simulation <- simulateOccJSDMData(
  list_datasettings = survey_settings,
  list_params = observation_settings,
  list_jsdmParams = ecological_settings,
  model = "two_stage"
)

new_survey_data <- new_simulation$data_list
new_known_truth <- new_simulation$true_params

# Label the truth matrices explicitly, as in the saved teaching dataset.
site_ids <- as.character(seq_len(n_sites))
sample_ids <- as.character(seq_len(n_field_samples))
species_ids <- colnames(new_survey_data$OTU)

dimnames(new_known_truth$z_true) <- list(site_ids, species_ids)
dimnames(new_known_truth$w_true) <- list(sample_ids, species_ids)
dimnames(new_known_truth$jsdmParams_true$eta) <- list(site_ids, species_ids)
```

The last block labels the truth matrices with site, sample and species IDs, as in the saved dataset, because the joins below match truth to observations by those identifiers.

If you change a setting, the old fitted results no longer belong to your new dataset: fit it again before comparing estimates with truth. Simply replacing a covariate column after simulation would also break the match between observations and their generating model. The appendix records the package revision, seed and R version needed to reproduce the saved simulation exactly.

## Understand the three input objects

occJSDM receives three tables, and your own data must take the same form: `info` describes each PCR reaction, `OTU` holds its read counts, and `traits` describes each species.

```{r inspect-inputs}
observation_info <- as_tibble(survey_data$info)
read_counts <- as_tibble(survey_data$OTU)

species_traits <- survey_data$traits |>
  as.data.frame() |>
  rownames_to_column("species") |>
  as_tibble()

observation_info |>
  slice_head(n = 6) |>
  knitr::kable(digits = 2)

read_counts |>
  slice_head(n = 6) |>
  knitr::kable()

species_traits |>
  knitr::kable(digits = 3)
```

- `info` has one row per PCR reaction: its `Site`, its field `Sample` and its `Primer`, followed by the covariates. Columns beginning `X_psi.` are occupancy covariates, `X_theta` is the collection covariate, and `Xs.1` and `Xs.2` are the site coordinates.
- `OTU` has one row per PCR reaction, in the same order as `info`, and one read-count column per species. OTU stands for operational taxonomic unit: these columns correspond to the OTU table from your own bioinformatics. Read counts can be supplied as they are, because `runOccJSDM()` converts each count to a detection or non-detection at its `threshold`. OTU_1, for example, is a simulated species label.
- `traits` has one row per species, with two simulated trait measurements. Its row names must match the column names of `OTU`, because `runOccJSDM()` matches traits to species by name.

The first six rows are six PCRs from the same sample and primer. Site conditions repeat because several samples and PCRs belong to the same site. Collection conditions repeat across PCRs from the same field sample.

Rows of `info` and `OTU` must stay paired. Sorting either one on its own would associate read counts with the wrong sample. When we join other tables below, we use explicit site or species identifiers.

## Follow the observations back to their known source

Because the simulation kept the truth, we can label every PCR with what actually produced its result. That shows the false positives and false negatives that the model will have to tell apart without these labels. Make a table with one row per **species and PCR reaction**: put the already aligned `info` and read-count tables side by side, and number the PCR replicates within each sample and primer. Then `pivot_longer()` puts species names and their read counts into two columns.

```{r make-observation-table}
observed_pcrs <- bind_cols(observation_info, read_counts) |>
  group_by(Site, Sample, Primer) |>
  mutate(PCR = row_number()) |>
  ungroup() |>
  pivot_longer(
    cols = all_of(colnames(survey_data$OTU)),
    names_to = "species",
    values_to = "reads"
  ) |>
  mutate(
    Site = as.character(Site),
    Sample = as.character(Sample),
    positive = as.integer(reads >= 1)
  )
```

The new table has `r format(nrow(observed_pcrs), big.mark = ",")` rows: `r format(n_pcr_observations, big.mark = ",")` PCR reactions for each of `r n_distinct(observed_pcrs$species)` species. `positive` is one for at least one read and zero for no reads, which matches the `threshold = 1` of Lesson 1's fits. A missing read count would remain missing; this survey has `r if (anyNA(observed_pcrs$reads)) sum(is.na(observed_pcrs$reads)) else "none"`. A PCR whose result was lost would be `NA`, whereas a field sample that was never analysed is represented by absent rows, as "Unequal numbers of field samples" shows. Species are columns in the matrix used by the fitter and rows in this table used for exploration; these are two representations of the same observations.

Now turn the true site and sample states into tables. The saved matrices have site or sample IDs as row names and species IDs as column names. `rownames_to_column()` makes the row identifier an explicit column.

```{r join-observations-to-truth}
presence_truth <- known_truth$z_true |>
  as.data.frame() |>
  rownames_to_column("Site") |>
  pivot_longer(cols = -Site, names_to = "species", values_to = "presence")

sample_truth <- known_truth$w_true |>
  as.data.frame() |>
  rownames_to_column("Sample") |>
  pivot_longer(cols = -Sample, names_to = "species", values_to = "dna_in_sample")

observations_with_truth <- observed_pcrs |>
  left_join(presence_truth, by = c("Site", "species")) |>
  left_join(sample_truth, by = c("Sample", "species")) |>
  mutate(
    source = case_when(
      is.na(positive) ~ "Missing",
      positive == 0 & presence == 0 ~ "True negative",
      positive == 0 ~ "Missed detection",
      dna_in_sample == 0 ~ "Laboratory false positive",
      presence == 0 ~ "Field-stage false positive",
      TRUE ~ "True detection"
    )
  )

weak_case <- lesson$cases |>
  filter(case == "Weak true detection")

weak_pcrs <- observations_with_truth |>
  filter(
    species == weak_case$species,
    Site == as.character(weak_case$Site),
    Sample == as.character(weak_case$Sample)
  ) |>
  select(Site, Sample, Primer, PCR, reads, positive, presence, dna_in_sample, source)

knitr::kable(weak_pcrs)
```

`left_join()` attaches the true states by site, sample and species identity, so the tables need not share a row order. Sample IDs are unique across the whole survey in this dataset, so sample plus species identifies a sample state. The order of `case_when()` matters. A negative PCR is a **true negative** if the species does not occupy the site and a **missed detection** if it does. A positive in a sample without DNA is a **laboratory false positive**, contamination in the laboratory, even if the species happened to occupy the site. A positive from DNA in an unoccupied site's sample is a **field-stage false positive**, contamination in the field. Any other positive is a **true detection**.

The table shows Lesson 1's weak true-detection example, `r weak_case$species` in Sample `r weak_case$Sample` at Site `r weak_case$Site`. The species occupies the site and its DNA is in the sample: `presence` and `dna_in_sample` are both 1. Yet only `r sum(weak_pcrs$positive)` of the `r nrow(weak_pcrs)` PCRs are positive, with `r knitr::combine_words(weak_pcrs$reads[weak_pcrs$positive == 1])` reads, from primer `r knitr::combine_words(unique(weak_pcrs$Primer[weak_pcrs$positive == 1]))`. The other `r sum(weak_pcrs$positive == 0)` are missed detections. That is what makes the case weak. A real survey would see a sample with only a few positives, which could come from DNA that the PCRs often miss or from contamination, and Lesson 1 asks which explanation the model favours.

These are **known source labels supplied by the simulation**, not classifications produced by the model. We do not pass `observations_with_truth` to the fitter. `lesson$cases` records the four detection examples that Lesson 1 selected before fitting, and Lesson 1 compares the fitted probabilities for this sample with these states.

Across the whole survey, how common is each outcome? Count the PCRs with each label, with the median read count of each.

```{r count-sources}
source_counts <- observations_with_truth |>
  group_by(source) |>
  summarise(pcrs = n(), median_reads = median(reads), .groups = "drop") |>
  arrange(desc(pcrs))

source_counts

pcrs_with <- setNames(source_counts$pcrs, source_counts$source)
```

Each row is one kind of outcome, `pcrs` counts the species-PCR combinations with it, and `median_reads` is their median read count. Most combinations are negative: `r format(pcrs_with[["True negative"]], big.mark = ",")` true negatives at sites the species does not occupy and `r format(pcrs_with[["Missed detection"]], big.mark = ",")` missed detections at sites it does. A missed detection here is any negative PCR at an occupied site. It includes the PCRs of samples whose DNA was never collected (`r format(sum(observations_with_truth$source == "Missed detection" & observations_with_truth$dna_in_sample == 0), big.mark = ",")` of them), as well as PCRs that missed DNA in the sample, so missed detections outnumber the `r format(pcrs_with[["True detection"]], big.mark = ",")` true detections. Of the `r format(sum(observations_with_truth$positive), big.mark = ",")` positive PCRs, `r format(pcrs_with[["Laboratory false positive"]], big.mark = ",")` are laboratory false positives and `r format(pcrs_with[["Field-stage false positive"]], big.mark = ",")` field-stage false positives, together `r format_percent((pcrs_with[["Laboratory false positive"]] + pcrs_with[["Field-stage false positive"]]) / sum(observations_with_truth$positive))` of all positives. The read counts follow the settings. Laboratory false positives have a median of `r source_counts$median_reads[source_counts$source == "Laboratory false positive"]` reads, against `r source_counts$median_reads[source_counts$source == "True detection"]` for true detections. Field-stage false positives have `r source_counts$median_reads[source_counts$source == "Field-stage false positive"]`, as many as true detections, because their samples really contain the species' DNA. A low read count can therefore hint at a laboratory false positive, but in this simulation nothing in the reads reveals contamination in the field.

## Put the sites and environmental conditions on a map

`distinct()` keeps one copy of each site's information. For plotting, `east` and `north` are arbitrary coordinates between zero and one, not longitude, latitude or kilometres.

```{r prepare-site-table}
site_covariates <- observation_info |>
  transmute(
    Site = as.character(Site),
    east = Xs.1,
    north = Xs.2,
    environment_1 = X_psi.EnvCov.1,
    environment_2 = X_psi.EnvCov.2
  ) |>
  distinct()

site_covariates |>
  slice_head(n = 6) |>
  knitr::kable(digits = 3)
```

The map marks the sites of the four detection examples in `lesson$cases`.

```{r sampling-locations, fig.height=4.5, fig.cap="The 100 simulated sampling sites. The highlighted sites hold the four detection examples used in Lesson 1; the laboratory false-positive and field-stage false-positive examples share Site 1. They were selected from known truth and observed detection patterns, before examining fitted probabilities."}
example_sites <- lesson$cases |>
  distinct(Site) |>
  mutate(Site = as.character(Site)) |>
  left_join(site_covariates, by = "Site")

ggplot(site_covariates, aes(x = east, y = north)) +
  geom_point(colour = "grey45", size = 2) +
  geom_point(data = example_sites, colour = "#D55E00", size = 3) +
  geom_text(data = example_sites, aes(label = Site), nudge_y = 0.035) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "East coordinate (arbitrary units)", y = "North coordinate (arbitrary units)") +
  theme_minimal(base_size = 12)
```

Next, `pivot_longer()` converts the two environmental columns into one value column and a label saying which covariate it belongs to, so that each covariate can have its own panel.

```{r environmental-maps, fig.height=4.5, fig.cap="Environmental values at the actual sampled sites. In this first dataset, neighbouring sites do not systematically share similar conditions. Both panels use the same arbitrary measurement scale; neither represents a real temperature or moisture measurement."}
environment_long <- site_covariates |>
  pivot_longer(
    cols = starts_with("environment_"),
    names_to = "covariate",
    values_to = "value"
  )

ggplot(environment_long, aes(x = east, y = north, colour = value)) +
  geom_point(size = 2.5) +
  facet_wrap(~ covariate) +
  scale_colour_viridis_c(name = "Covariate value") +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_equal() +
  labs(x = "East coordinate", y = "North coordinate") +
  theme_minimal(base_size = 12)
```

The covariate values are in arbitrary units, from `r round(min(environment_long$value), 1)` to `r round(max(environment_long$value), 1)` across both covariates. The patchiness is intentional in this dataset: the simulator drew environmental values independently of coordinates. We show points instead of interpolating a smooth surface that was never simulated. The spatial lesson will introduce smooth environmental gradients, so that geography becomes informative about habitat conditions.

## Map probability, actual occurrence and detection separately

Consider OTU_1 and OTU_10, the species used in Lesson 1's detection examples. The settings above make them the hardest and the easiest species to detect, so their maps show what detectability does. The species are chosen to connect the lessons, not because their fitted maps look especially good. First calculate whether there was any positive PCR for each species at each site. This deliberately simple detection rule is an observation summary, not an occupancy model. It misses occupied sites where no PCR found the species' DNA, and it counts unoccupied sites where contamination produced a positive, so it can understate or overstate where a species occurs.

```{r prepare-species-maps}
site_detections <- observations_with_truth |>
  group_by(Site, species) |>
  summarise(
    any_positive = if (all(is.na(positive))) {
      NA_real_
    } else {
      as.numeric(any(positive == 1, na.rm = TRUE))
    },
    .groups = "drop"
  ) |>
  mutate(Site = as.character(Site))

# Convert the true occupancy scores to probabilities, then to a long table.
occupancy_truth <- known_truth$jsdmParams_true$eta |>
  plogis() |>
  as.data.frame() |>
  rownames_to_column("Site") |>
  pivot_longer(cols = -Site, names_to = "species", values_to = "probability")

species_truth <- occupancy_truth |>
  filter(species %in% c("OTU_1", "OTU_10")) |>
  left_join(presence_truth, by = c("Site", "species")) |>
  left_join(site_detections, by = c("Site", "species")) |>
  left_join(site_covariates, by = "Site")

truth_map_data <- species_truth |>
  pivot_longer(
    cols = c(probability, presence, any_positive),
    names_to = "quantity",
    values_to = "value"
  ) |>
  mutate(
    quantity = factor(
      quantity,
      levels = c("probability", "presence", "any_positive"),
      labels = c(
        "True occupancy probability", "Actual presence (0 or 1)",
        "Any positive PCR (0 or 1)"
      )
    )
  )
```

The explicit missing-data check prevents an unobserved site from being treated as a negative. `eta` is each site's true occupancy score for each species: the sum of the baseline, environmental responses and hidden-factor effects, on the logit scale. `plogis()`, the inverse logit, converts it to an occupancy probability. `factor()` sets the order of panels.

```{r truth-and-observation-maps, fig.width=9, fig.height=6, fig.cap=paste0("Left: generating occupancy probability. Middle: the actual presence/absence drawn using that probability. Right: whether any of the ", samples_per_site * n_primers * pcrs_per_primer, " PCRs at that site was positive. The middle and right panels contain only zeros and ones; their differences expose missed occurrences and false detections. All values come from the matching simulation; no fitted estimates are shown here.")}
ggplot(truth_map_data, aes(x = east, y = north, colour = value)) +
  geom_point(size = 2) +
  facet_grid(species ~ quantity) +
  scale_colour_viridis_c(limits = c(0, 1), name = "Probability / state") +
  scale_x_continuous(breaks = c(0, 0.5, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  coord_equal() +
  labs(x = "East coordinate", y = "North coordinate") +
  theme_minimal(base_size = 11)
```

Read one row of maps from left to right. A high-probability site need not be occupied in this particular realization. An occupied site can have no positive PCRs. An unoccupied site can have a positive PCR, from contamination in the field or a laboratory false positive. To see how often each happens, count the sites in each combination of true presence and any positive PCR. Then, at the occupied sites, check whether any field sample collected the species' DNA.

```{r presence-by-detection}
species_truth |>
  count(species, presence, any_positive)

occupied_sites <- observations_with_truth |>
  filter(species %in% c("OTU_1", "OTU_10"), presence == 1) |>
  group_by(species, Site) |>
  summarise(
    dna_collected = any(dna_in_sample == 1),
    any_positive = any(positive == 1),
    .groups = "drop"
  )

occupied_sites |>
  count(species, dna_collected, any_positive)
```

In the first table, `presence` is the true site state, `any_positive` is the simple rule and `n` counts sites. OTU_1 occupies `r sum(species_truth$species == "OTU_1" & species_truth$presence == 1)` sites and OTU_10 only `r sum(species_truth$species == "OTU_10" & species_truth$presence == 1)`, so the two species fail the rule in opposite ways.

OTU_1 is hard to collect and detect, and the second table shows where that bites. At `r with(occupied_sites, sum(species == "OTU_1" & !dna_collected))` of its occupied sites, no field sample collected its DNA. The rule still shows a positive at `r with(occupied_sites, sum(species == "OTU_1" & !dna_collected & any_positive))` of them, but only through laboratory false positives. The remaining `r with(species_truth, sum(species == "OTU_1" & presence == 1 & any_positive == 0))` are the occupied sites that the rule misses. OTU_1 is absent from only `r with(species_truth, sum(species == "OTU_1" & presence == 0))` sites, and `r with(species_truth, sum(species == "OTU_1" & presence == 0 & any_positive == 1))` of them have a positive PCR. Its map of positives therefore looks much like its map of presence, but partly for the wrong reasons.

OTU_10 is easy to collect and detect. Its DNA reached a field sample at `r with(occupied_sites, if (all(dna_collected[species == "OTU_10"])) paste("every one of its", sum(species == "OTU_10")) else paste(sum(species == "OTU_10" & dna_collected), "of its", sum(species == "OTU_10")))` occupied sites, and the rule misses `r if (with(species_truth, any(species == "OTU_10" & presence == 1 & any_positive == 0))) with(species_truth, sum(species == "OTU_10" & presence == 1 & any_positive == 0)) else "none"` of them. Its errors run the other way. It is absent from `r with(species_truth, sum(species == "OTU_10" & presence == 0))` sites, and `r with(species_truth, sum(species == "OTU_10" & presence == 0 & any_positive == 1))` of them have a positive PCR, `r with(observations_with_truth, n_distinct(Site[species == "OTU_10" & source == "Field-stage false positive"]))` of them with a field-stage false positive. With `r samples_per_site * n_primers * pcrs_per_primer` PCRs at each site, even a few percent chance of a laboratory false positive in each PCR makes at least one stray positive likely, and OTU_10 also has the highest field contamination rate. The rule therefore places OTU_10 at `r with(species_truth, sum(species == "OTU_10" & any_positive == 1))` sites, against `r sum(species_truth$species == "OTU_10" & species_truth$presence == 1)` truly occupied. Lesson 1 asks how much of this uncertainty the fitted model can resolve.

## Unequal numbers of field samples

Real surveys may lose a field sample. Here we remove one **whole field sample at each of three distinct sites**, keeping every PCR row of all remaining samples. The choice uses seed 3947, declared before fitting so that the removed samples could not be chosen after seeing any results. It does not use the species' states, PCR detections or model results.

```{r unbalanced-select-samples}
sample_keys <- survey_data$info |>
  distinct(Site, Sample) |>
  arrange(Site, Sample)

set.seed(3947)
removed_sites <- sample(sort(unique(sample_keys$Site)), size = 3)
removed_samples <- sample_keys |>
  filter(Site %in% removed_sites) |>
  group_by(Site) |>
  slice_sample(n = 1) |>
  ungroup() |>
  arrange(Site)

removed_samples
```

This removes `r knitr::combine_words(paste("Sample", removed_samples$Sample, "from Site", removed_samples$Site), oxford_comma = FALSE)`. `group_by(Site)` makes `slice_sample(n = 1)` choose exactly one sample within each selected site. We keep the original global sample IDs, so the remaining IDs are no longer consecutive. occJSDM accepts that: [Lesson 1's fit of this survey](occJSDM-lesson-1.md#fit-a-survey-with-unequal-replication) ran with the original IDs.

The metadata and OTU matrix describe the same PCR rows. `anti_join()` drops metadata rows whose `(Site, Sample)` key is in the removal table. Keeping their original row numbers lets us apply **the identical selection to both tables**. The chunk then counts the samples at each site and the PCRs in each sample and primer, and `stopifnot()` checks the reduced survey's structure. It confirms the expected numbers of rows and samples, that every site remains, that every remaining sample has both primers and six PCRs, and that the species and traits are unchanged; it stops with an error if any check fails. Run checks like these after every subset of your own data.

```{r unbalanced-paired-rows}
retained_rows <- survey_data$info |>
  mutate(original_row = row_number()) |>
  anti_join(removed_samples, by = c("Site", "Sample")) |>
  pull(original_row)

unbalanced_data <- survey_data
unbalanced_data$info <- survey_data$info[retained_rows, , drop = FALSE]
unbalanced_data$OTU <- survey_data$OTU[retained_rows, , drop = FALSE]

samples_per_site_unbalanced <- unbalanced_data$info |>
  distinct(Site, Sample) |>
  count(Site, name = "samples")

pcrs_per_primer_unbalanced <- unbalanced_data$info |>
  count(Site, Sample, Primer, name = "PCRs")

stopifnot(
  nrow(unbalanced_data$info) == 3564,
  nrow(unbalanced_data$OTU) == 3564,
  n_distinct(unbalanced_data$info$Sample) == 297,
  nrow(samples_per_site_unbalanced) == 100,
  all(samples_per_site_unbalanced$samples >= 1),
  nrow(pcrs_per_primer_unbalanced) == 297 * 2,
  all(pcrs_per_primer_unbalanced$PCRs == 6),
  identical(colnames(unbalanced_data$OTU), colnames(survey_data$OTU)),
  identical(unbalanced_data$traits, survey_data$traits)
)

samples_per_site_unbalanced |>
  count(samples, name = "sites")
```

There are `r sum(samples_per_site_unbalanced$samples == samples_per_site)` sites with `r samples_per_site` samples and `r sum(samples_per_site_unbalanced$samples == samples_per_site - 1)` sites with `r samples_per_site - 1`. We removed `r nrow(removed_samples)` samples and `r nrow(survey_data$info) - nrow(unbalanced_data$info)` PCR rows: `r n_distinct(unbalanced_data$info$Sample)` field samples and `r format(nrow(unbalanced_data$info), big.mark = ",")` PCR rows remain, still covering all `r nrow(samples_per_site_unbalanced)` sites and `r ncol(unbalanced_data$OTU)` species. Every retained sample has both primers and six PCRs per primer. No retained read count or covariate changes.

A lost sample is represented by **absent rows**. It is not an observed sample with zero reads, and it is not a collection of `NA` PCR results. The original `known_truth` still describes all `r nrow(known_truth$z_true)` sites, all `r ncol(known_truth$z_true)` species and all `r nrow(known_truth$w_true)` simulated samples, including those we removed from the observed survey. Lesson 1 fits this reduced survey and compares the resulting probabilities with those same truths.

Unequal replication at every level is accepted by the fitter: different numbers of field samples per site, of primers per sample, and of PCRs per sample and primer. `runOccJSDM()` counts the rows in each site, sample and primer block, so nothing needs declaring. Besides Lesson 1's fit of this survey, the package's tests fit a survey in which some samples lack a primer and blocks have one to three PCRs. On your own survey, drop the rows of whatever was lost, keep the identifiers of what remains, and judge the result by the fit. `plotCumulativeSpeciesDetections()` is no guide here: it takes a single number of samples per site and of PCRs, so it describes a balanced design.

## Take the right objects into Lesson 1

The main two-stage fit receives **`survey_data` only**; the unbalanced extension receives **`unbalanced_data` only**. It does not receive the true occupancy probabilities, presence states, sample states or source labels. Lesson 1 also includes a deliberately labelled perfect-observation control: a fit given the true presence/absence matrix instead of the PCRs, which shows how well the model could do if detection were perfect. That control still has to estimate the probabilities that generated those binary states.

[The spatial lesson](occJSDM-lesson-2.md) keeps this observation process and changes the ecological simulation.

The appendix below records the package revision, seed and software environment of the saved simulation, and how the lesson's code for the unequal survey is checked.

Continue to [Lesson 1: Fit the model and compare its answers with truth](occJSDM-lesson-1.md).

## Appendix: evidence and reproduction

The original simulation used occJSDM revision **`r substr(lesson$input$source_commit, 1, 7)`**, seed **`r lesson$input$seed`** and **`r lesson$manifests$default$session$R.version$version.string`**. Exact reproduction requires the recorded code and software environment as well as the seed, because a different package or R version can turn the same seed into different random draws. To install that revision, pass it as the `ref` argument of `remotes::install_github("AlexDiana/occJSDM")`. The package versions are recorded in `lesson$manifests[["default"]]$session`. Instructions for regenerating the saved bundle are in [the lesson build README](https://github.com/AlexDiana/occJSDM/blob/main/dev/simstudy/vignette-lesson/README.md).

The two chunks that prepare the unequal survey, `unbalanced-select-samples` and `unbalanced-paired-rows`, are checked by `dev/simstudy/vignette-lesson/unbalanced-verify.R`. It runs them exactly as displayed here and confirms that they produce the reduced survey of Lesson 1's saved fit, with every truth unchanged.
