---
title: "occJSDM: quickstart"
output:
  rmarkdown::html_vignette:
    toc: true
    css: teaching.css
vignette: >
  %\VignetteIndexEntry{occJSDM: quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, comment = "#>")

source("lesson-links.R")
```

`occJSDM` estimates which species occupy which sites from eDNA surveys. It allows for false negatives (from imperfect field collection, imperfect PCR detection or both) and for false positives (from contamination in the field, in the lab or both). It models the species jointly: their responses to the environment, the role of their traits, and associations between species that the covariates do not explain, optionally with a spatial field. It can also analyse ordinary presence/absence surveys, in which each species is recorded as present or absent at each site and the records are assumed to be free of detection error. In that case `occJSDM` fits only a joint species distribution model (JSDM).

This quickstart shows how the input data must be structured, makes a fitting call and takes a first look at the output, using the example data shipped with the package.

## The example data

`sampledata` is a simulated two-stage eDNA survey of 10 species at 100 sites, stored as an R list. Each site has three field samples, and each sample was tested with three primers, twice each.

```{r data}
library(occJSDM)

sampledata <- sampledata # this loads the package's sampledata into your environment
str(sampledata, max.level = 1)
head(sampledata$info)
head(sampledata$OTU)
head(sampledata$traits)
```

It has three parts:

- `info` has one row per PCR reaction: the `Site`, the field `Sample` the DNA came from, the `Primer`, and the covariates. Columns beginning `X_psi.` are occupancy covariates, columns beginning `X_theta.` are collection covariates, and `Xs.1` and `Xs.2` are the site coordinates.
- `OTU` has one column per species and one row for each row of `info`, holding that species' read count in that reaction. Read counts can be supplied as they are, without converting the OTU table first: `runOccJSDM()` converts each count to a detection or non-detection, and a PCR reaction with at least `threshold` reads counts as a detection.
- `traits` has one row per species.

`runOccJSDM()` uses the structure of `info` to choose which model to fit. Here each sample has several rows, one per primer and PCR replicate, so it fits the two-stage occupancy model. If instead `info` had only one row per site, with no sample or PCR replicates, `runOccJSDM()` would skip the detection stages and fit a JSDM to the observed presence/absence.

## Fit the model

```{r fit, eval=FALSE}
set.seed(20261007)

fit <- runOccJSDM(
  data = sampledata,
  listParams = list(n_factors = 2),
  threshold = 1,
  occCovariates = c("X_psi.EnvCov.1", "X_psi.EnvCov.2"),
  collCovariates = c("X_theta.1", "X_theta.2"),
  spatCovariates = c("Xs.1", "Xs.2"),
  MCMCparams = list(nchain = 2, nburn = 5000, niter = 5000, nthin = 1),
  summarisedLatentPresences = TRUE
)
```

- `threshold`: a reaction with at least this many reads counts as a detection.
- `occCovariates`, `collCovariates` and `spatCovariates` name the columns of `info` to use for occupancy, for collection and for the spatial field. The species traits are taken from `sampledata$traits`.
  - The example includes the `spatCovariates` line to show how the full model is specified. The beta's spatial model is very underpowered, though, so we recommend leaving that line out of your own fits for now.
- `n_factors` is the number of latent factors that capture associations between species.
- `MCMCparams` sets the number of chains, the burn-in and kept iterations, and the thinning.
- The priors have defaults, which `listPriors` changes. `?runOccJSDM` describes every argument.

This chunk is not run while the vignette is built, because the fit takes several minutes. The package ships its result as `sampleresults`, which the rest of this guide uses. The beta example was refitted with the current package using the seed above, the default priors and one requested computation thread. A fixed seed reproduces the fit on the same software and platform; small numerical differences can occur on other platforms.

```{r load-fit}
fit <- sampleresults
```

## Check that the chains agree

Before reading any estimate, check that the chains have converged on the same answer.

```{r diagnostics}
diagnostics <- returnConvergenceDiagnostics(fit)
summary(diagnostics$rhat)
summary(diagnostics$ess)
head(diagnostics[order(-diagnostics$rhat), c("param", "label1", "label2", "rhat", "ess")])
```

`rhat` compares the chains with each other: values close to 1 mean they agree, and larger values call for longer runs or a closer look at the parameters concerned. `ess` is the effective number of independent draws behind each estimate. The table covers the occupancy and detection coefficients; `plotTraceplot()` shows the chains themselves.

The shipped run still has 35 of its 130 reported coefficients above `rhat = 1.01` or below `ess = 400`. Its largest reported Rhat is about 1.083, for the field false-positive rate of `OTU_3`, and fitting also raised warnings about the spatial coefficients. The plots below demonstrate the output functions. Before interpreting flagged estimates in a scientific analysis, run longer chains and check them again; more iterations do not guarantee that the chains will agree.

## Look at the results

Each species' baseline occupancy probability, with its credible interval:

```{r occupancy, fig.width=7, fig.height=4}
plotOccupancyRates(fit, idx_species = 1:5)
```

Each species' probability, for each primer, that a PCR reaction detects its DNA when that DNA is present in the sample:

```{r detection, fig.width=7, fig.height=4}
plotDetectionRates(fit, idx_species = 1:5)
```

The package's other `plot...()` and `return...()` functions cover covariate effects, collection and false-positive rates, species associations, ordination, variation partitioning and prediction at new sites. Each has its own help page.

## Where to go next

- `?runOccJSDM` for every fitting argument, including the priors.
- `vignette("simulateOccJSDMData", package = "occJSDM")` to simulate a survey whose truth you know, so that you can compare the model's answers with it. The truth behind `sampledata` is not stored with it.
- The [Known limitations](https://github.com/AlexDiana/occJSDM#known-limitations) section of the README, before relying on the estimates.

The detailed teaching lessons are first drafts only and are awaiting a full line-by-line review.
