---
title: "occJSDM"
output: 
  rmarkdown::html_vignette:
    fig.width: 7
    fig.height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{occJSDM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  comment = "#>"
)

options(warning = -1)
```

```{r setup, include = FALSE}
# library(occPlus)
devtools::load_all() # change to library(occPlus) when package finished
```

The model can be run using the `runOccPlus` function.

The arguments of the function are:

- `data`: which we explain below;
- `listParams`: a list of the number of ordination factors used for
  computing occupancy probabilities and of the number of unobserved
  traits;
- `threshold`: ≥ this number of reads is counted as a detection (default
  = 1);
- `occCovariates`: vectors of covariates predicting occupancy
  probabilities;
- `collCovariates`: vectors of covariates predicting eDNA collection (=
  detection) probabilities;
- `spatCovariates`: vectors of covariates (e.g. X, Y) used to construct
  a spatial random field capturing residual spatial structure in
  occupancy that is not explained by occCovariates;
- `traitsMatrix`: a matrix of species traits (rows = species, columns =
  traits) used to explain how species' occupancy responses to
  `occCovariates` vary as a function of their traits;
- `MCMCparams`: a list specifying the MCMC settings, including `nchain`
  (number of chains), `nburn` (number of burn-in iterations to discard),
  `niter` (number of post-burn-in iterations to keep), and `nthin`
  (thinning interval, i.e. how many iterations to skip between saved
  samples).

## Data preparation

The data object is a list of two elements: `info` and `OTU`.

```{r}
data <- data
head(data$info)
```

`data$info` is a data.frame with $N$ rows, where $N$ is the total number
of samples analyzed and columns:

- `Site`: name (or index) of the site in which this replicate was
  collected;
- `Sample`: name (or index) of the sample from which this replicate was
  analyzed;
- `Primer`: name (or index) of the primer used in this replicate;
- `X_psi.n` are the occupancy covariates (e.g. elevation);
- `X_theta.n` are the eDNA collection covariates (e.g. water volume per
  sample).

In this dataset, there are $100$ sites and, in the balanced case, $m=3$
sample replicates per site and $L=3$ primers. [From this data
structure]{.underline}, [occJSDM can infer that there are $K=2$ PCR
replicates per sample, so a fully balanced dataset would have had
$3 \times 3 \times 2 = 1800$ rows (i.e. observations). However, $3$ of
the $300$ samples (one each at three different sites) are dropped to
simulate missing observations, leaving $297$ samples and $1{,}782$ total
replicates. This illustrates that occJSDM can work with unbalanced study
designs, such as sites with a missing samples.]{.underline}

We can see this imbalance directly by counting the number of rows per
site: most sites have $18$ rows ($3$ samples $\times$ $3$ primers
$\times$ $2$ PCR replicates), but the three affected sites have only
$12$.

```{r}
data$info |> count(Site)
```

```{r}
head(data$OTU)
```

`data$OTU` is a matrix of dimension $N \times S$, where $N$ is again the
number of samples (in the same order as `data$info`), and $S$ is the
number of species. Each element in the matrix is the number of reads of
the species in that replicate, [which will be converted to
detection/non-detection (i.e. 1/0)]{.underline}.

## Model fitting

Next, we can run the model by typing (leaving out `spatCovariates` and
`traitsMatrix` for now)

```{r, eval = FALSE}
fitmodel  <- runOccPlus(data,
                        listParams = list(n_factors = 2, 
                                          gt = 0),
                        threshold = 1,
                        occCovariates = c("X_psi.1", "X_psi.2"),
                        collCovariates = c("X_theta"),
                        MCMCparams = list(nchain = 2,
                                          nburn = 500,
                                          niter = 500)
                        );beepr::beep()
```

The names of the covariates in `occCovariates` and `collCovariates`
should match the column names in `data$info`.

## Output

There are several outputs we can obtain:

### Covariate coefficients

First, we provide two functions for returning collection and occupancy
covariate coefficients. The output is returned in the form (*niter* x
*numberOfCovariates* x *numberOfSpecies*).

```{r}
collCovariatesOutput <- returnCollectionCovariates(fitmodel)
str(collCovariatesOutput)
```

```{r}
occCovariatesOutput <- returnOccupancyCovariates(fitmodel)
str(occCovariatesOutput)
```

We provide two functions for plotting collection and occupancy covariate
coefficients:

```{r}
plotCollectionCovariates(fitmodel, idx_species = 1:10, 
                         covName = "X_theta")
```

```{r}
plotOccupancyCovariates(fitmodel, idx_species = 1:10, 
                        covName = "X_psi.1")

plotOccupancyCovariates(fitmodel, idx_species = 1:10, 
                        covName = "X_psi.2")
```

### Collection, Detection, and False-Positive rates

Second, we can plot collection, detection, and false-positive rates for
Stage 1 (eDNA collection in the field) and Stage 2 (eDNA detection in
the lab).

```{r Stage 1}
plotStage1FPRates(fitmodel, idx_species = 1:10) + ylim(c(0,1)) # ideally, these are low
plotCollectionRates(fitmodel, idx_species = 1:10) # ideally, these are high
```

```{r Stage 2}
plotStage2FPRates(fitmodel, idx_species = 1:10) + ylim(c(0,1)) # ideally, these are low
plotDetectionRates(fitmodel, idx_species = 1:10) # ideally, these are high. 
```

The Stage 2 detection rates are calculated separately for each primer
and species, which allows one to determine how well each primer
amplifies each species.

### Occupancy probabilities

To compute *conditional* occupancy probabilities at each site and for
each species, we use the following function. In occJSDM, the conditional
occupancy means that the occupancy probability is accounting for both
the covariates *and* the observed detection data at each site.

```{r}
condOccProbs <- computeConditionalOccupancyProbs(fitmodel)
str(condOccProbs)
```

Finally, to allow users to build their own diagnostics, we can return
baseline occupancy probabilities, which are the the logits of the
intercepts (i.e. the occupancy of each species when all covariates are
set to the normalised mean (= 0). The output is returned in the form
(*numberOfSpecies* X *niter* X *nchain*). Because covariates are set to
their mean values, there is no site information.

```{r}
baselineOccupancyRates <- returnBaselineOccupancyRates(fitmodel) 
str(baselineOccupancyRates)
```

### 

### Latent presences

To understand how occJSDM estimates occupancies as a function of

```{r}
latentPresences <- returnLatentPresences(fitmodel, idx_species = 1)
View(latentPresences)
```

We can display `latentPresences` as a `gt` table with a nested banding
scheme: the fill colour changes each time we move to a new `Site`, and
alternates between a light and a dark shade of that colour each time we
move to a new `Sample` within the site. This makes both the site and
sample boundaries visually clear across all rows.

```{r}
library(gt)

base_colors <- c("#4E79A7", "#F28E2B", "#59A14F", "#E15759")

# four lighten amounts (lightest -> darkest), indexed by combining
# sample_parity (coarse banding) with primer_parity (fine sub-shade)
shade_amounts <- c(0.78, 0.60, 0.45, 0.28)
shade_lookup <- sapply(shade_amounts, function(amt) colorspace::lighten(base_colors, amt))

latentPresences_banded <- latentPresences |>
  dplyr::mutate(
    # a new group id each time Site changes
    site_group   = cumsum(Site   != dplyr::lag(Site,   default = dplyr::first(Site))),
    # a new group id each time Sample changes (nested within Site)
    sample_group = cumsum(Sample != dplyr::lag(Sample, default = dplyr::first(Sample))),
    # a new group id each time Primer changes (nested within Sample)
    primer_group = cumsum(Primer != dplyr::lag(Primer, default = dplyr::first(Primer))),
    site_color_idx = (site_group %% length(base_colors)) + 1,
    sample_parity   = sample_group %% 2,
    primer_parity   = primer_group %% 2,
    # 2-bit code: coarse (Sample) then fine (Primer) sub-shade
    shade_level     = sample_parity * 2 + primer_parity + 1,
    row_color       = shade_lookup[cbind(site_color_idx, shade_level)]
  )

tbl <- latentPresences_banded |>
  dplyr::select(Site, Sample, Primer, OTU, MeanSitePresence,
                MeanSamplePresence, PredOccProb, CollectionProb, row_color) |>
  gt() |>
  fmt_number(columns = c(MeanSitePresence, MeanSamplePresence,
                         PredOccProb, CollectionProb), decimals = 2) |>
  cols_hide(row_color) |>
  tab_header(
    title = "Latent presence summary (species 1)",
    subtitle = "Colour: Site | Shade: Sample (coarse) x Primer (fine)"
  ) |>
  tab_options(
    container.height = px(600),
    container.overflow.y = "scroll"
  ) |>
  tab_source_note(
    source_note = gt::html(paste0(
      "<b>Colour</b> = Site &middot; <b>shade</b> = Sample (coarse) ",
      "within a Site &middot; <b>fine shade</b> = Primer within a Sample."
    ))
  )

for (col in unique(latentPresences_banded$row_color)) {
  tbl <- tbl |>
    tab_style(
      style = cell_fill(color = col),
      locations = cells_body(rows = latentPresences_banded$row_color == col)
    )
}

tbl
```

### Residual correlation matrix

We can plot the posterior median of the *residual* correlation matrix,
after accounting for occupancy covariates (positive in red and negative
in blue). X's mark non-significant correlations.

```{r}
plotResidualCorrelationMatrix(fitmodel, idx_species = 1:10)
```
