---
title: "Lesson 3: Understand the model's outputs by comparing them with truth"
output:
  rmarkdown::html_vignette:
    toc: true
    css: teaching.css
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Lesson 3: Understand the model's outputs by comparing them with truth}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: none
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, collapse = FALSE, comment = "#>",
  fig.width = 9, fig.height = 5, dpi = 150, dev = "png"
)

lesson_link <- function(name) {
  is_html <- knitr::pandoc_to() %in% c("html", "html4", "html5")
  paste0(name, if (is_html) ".html" else ".md")
}
```

## What this lesson answers

An environmental effect can be real in the simulation and still be estimated imprecisely. A fitted curve can look convincing and still miss the truth. This lesson shows how to distinguish those situations.

It replaces the output tour in the original occJSDM vignette. Start with [Lesson 1](`r lesson_link("occJSDM-first-lesson")`) for fitting and false-positive interpretation; [Lesson 0](`r lesson_link("occJSDM-lesson-0")`) explains the simulation. You do not need the planned spatial Lesson 2 first. Here we use the same **non-spatial** community: 100 sites, 10 species, two measured environmental covariates, two measured traits, two field samples per site, two primers and six PCR replicates per primer.

We compare two existing fits of that community: one given the actual presence/absence matrix (**perfect observation**) and one given the PCR observations. The diagnostics section also revisits Lesson 1's longer alternative-prior fit. This is not a before/after comparison of software versions. The fits use the verified model source recorded in Lesson 1; that source matches the code on main when this lesson was prepared. No new MCMC was needed to make these figures.

All teaching code is visible. Run chunks with `vignettes` as the working directory, or knit this file. Figures and tables use compact saved summaries. Optional chunks labelled `eval=FALSE` explain how to obtain the underlying outputs from a full fit, without starting a long fit while knitting.

```{r load-results, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

lesson <- readRDS("teaching-data/nonspatial-lesson.rds")
outputs <- readRDS("teaching-data/output-lesson.rds")
diagnostic_examples <- readRDS("teaching-data/diagnostics-lesson.rds")
known_truth <- lesson$input$sim$true_params

fit_labels <- c(
  perfect = "Perfect observation",
  default = "PCR observations"
)

environment_labels <- c(
  X_psi.EnvCov.1 = "Environmental gradient 1",
  X_psi.EnvCov.2 = "Environmental gradient 2"
)

species_order <- colnames(lesson$input$sim$data_list$OTU)

# This theme also works after occJSDM loads its ternary-plot dependency.
theme_set(ggtern::theme_bw(base_size = 12))
```

In the figures, **black crosses or lines show truth**. Blue points or lines show estimates, and blue intervals show posterior uncertainty. An interval is not a measurement of how far the estimate actually is from truth; simulation lets us check both separately.

## Do environmental effects really go undetected?

Each environmental coefficient describes how one species responds to one gradient, after accounting for the other model components. With ten species and two gradients, there are twenty coefficients.

The older `sampleresults` object currently shipped with the package has `r outputs$legacy_counts$excludes_zero[outputs$legacy_counts$block == "environment"]` of `r outputs$legacy_counts$total[outputs$legacy_counts$block == "environment"]` environmental intervals and `r outputs$legacy_counts$excludes_zero[outputs$legacy_counts$block == "trait"]` of `r outputs$legacy_counts$total[outputs$legacy_counts$block == "trait"]` trait intervals excluding zero. Thus that particular saved fit does not show an absence of all effects. Its generating truth is not stored with it, so we cannot tell from that object alone how accurately it recovered the original effects. The results below use the newer matching simulation and fits instead.

```{r summarize-effect-evidence}
effect_counts <- outputs$coefficients |>
  group_by(arm, block) |>
  summarise(
    effects = n(),
    intervals_excluding_zero = sum(excludes_zero),
    .groups = "drop"
  ) |>
  mutate(fit = unname(fit_labels[arm])) |>
  select(fit, block, effects, intervals_excluding_zero)

knitr::kable(effect_counts, caption = "How many 95% intervals exclude zero?")
```

Perfect observation identifies the direction of more environmental effects in this example. That makes ecological sense: it removes uncertainty about whether each species was present. It does not reveal the underlying occurrence probabilities or guarantee precise coefficients.

### Put the true coefficient beside its estimate

The public function returns an array: posterior draws by environmental covariate by species. A full fit is available after following the reproduction instructions at the end of this lesson.

```{r extract-environment-coefficients, eval=FALSE}
environment_draws <- occJSDM::returnOccupancyCovariates(fitmodel)

example_draws <- environment_draws[, "X_psi.EnvCov.1", "OTU_1"]

tibble(
  estimate = mean(example_draws),
  lower = quantile(example_draws, 0.025),
  upper = quantile(example_draws, 0.975)
)
```

For the complete comparison we use the same calculation for every coefficient. The exporter joins each summary to the generating coefficient by its species and covariate identifiers. The environmental predictor scale was checked by reconstructing the simulation's full ecological predictor from the fitted design matrix and the true parameters.

```{r environmental-coefficients, fig.height=8}
environment_effects <- outputs$coefficients |>
  filter(block == "Environment") |>
  mutate(species = factor(term, levels = species_order))

ggplot(environment_effects, aes(x = species)) +
  geom_hline(yintercept = 0, colour = "grey55", linetype = "dashed") +
  geom_pointrange(
    aes(y = estimate, ymin = lower, ymax = upper),
    colour = "#0072B2"
  ) +
  geom_point(aes(y = truth), shape = 4, size = 3, stroke = 1) +
  facet_grid(
    covariate ~ arm,
    labeller = labeller(covariate = environment_labels, arm = fit_labels)
  ) +
  labs(
    x = "Species",
    y = "Environmental effect on log-odds of occurrence",
    caption = "Black cross: true coefficient. Blue point: posterior mean. Bar: 95% interval."
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

A positive coefficient means occurrence becomes more likely along that gradient; a negative one means less likely. Zero means no direct response to that covariate in this model. These coefficients are changes in log-odds for a one-standard-deviation change in the environmental predictor, not percentage-point changes in occurrence probability.

Read each species in three steps: where is the black cross, how far is the blue point from it, and how wide is the blue interval? If the interval crosses zero, the fitted model has not clearly resolved the direction under this criterion. It does **not** establish that the environmental effect is absent. Conversely, excluding zero does not guarantee an accurate effect size.

## What does an effect mean for a species' distribution?

A response curve translates a coefficient into occurrence probabilities. Here we change one measured gradient, hold the other at its median, and set the hidden site-factor contribution to zero. Both the true and fitted curves use those same conditions.

These are **environmental response profiles**, not maps, fitted probabilities at particular surveyed sites, or predictions averaged over unknown conditions at a new site. Those quantities answer different questions. The public function used here, `returnOccupancyGradient()`, evaluates the zero-factor profile.

```{r extract-response-profile, eval=FALSE}
response_profile <- occJSDM::returnOccupancyGradient(
  fitmodel,
  covName = "X_psi.EnvCov.1",
  n_grid = 40
)
```

```{r response-profile-1, fig.height=7}
response_curves <- outputs$gradients |>
  filter(arm == "default") |>
  mutate(species = factor(species, levels = species_order))

plot_response <- function(covariate_name) {
  response_curves |>
    filter(covariate == covariate_name) |>
    ggplot(aes(x = x)) +
    geom_ribbon(aes(ymin = low, ymax = high), fill = "#0072B2", alpha = 0.2) +
    geom_line(aes(y = med), colour = "#0072B2", linewidth = 0.7) +
    geom_line(aes(y = truth), colour = "black", linetype = "dashed", linewidth = 0.7) +
    facet_wrap(~ species, ncol = 5) +
    scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
    labs(
      title = unname(environment_labels[covariate_name]),
      x = "Environmental value (standard deviations)",
      y = "Occurrence probability",
      caption = "Black dashed: truth. Blue: posterior median and 95% interval. Hidden site factors set to zero."
    )
}

plot_response("X_psi.EnvCov.1")
```

```{r response-profile-2, fig.height=7}
plot_response("X_psi.EnvCov.2")
```

Notice that the same coefficient can produce very different probability changes depending on where the species starts on the vertical axis. An effect near a baseline probability of 50% has more room to move the probability than the same log-odds change near 0% or 100%. The curves make that easier to see than the coefficient plot alone.

`returnOccupancyRates()` similarly describes the probability when the standardized measured predictors and the hidden factors are all zero. It is not the average occupancy across the landscape. Here are the posterior means of that baseline probability and their matching true values.

```{r baseline-probabilities}
outputs$baseline |>
  filter(arm == "default") |>
  select(species, truth, estimate, lower, upper) |>
  mutate(across(c(truth, estimate, lower, upper), ~ scales::percent(.x, accuracy = 0.1))) |>
  knitr::kable(caption = "Baseline probabilities: all predictor contributions set to zero.")
```

## Traits ask a harder, different question

An environmental coefficient asks, for example, whether a particular species becomes more likely along a gradient. A trait effect asks whether **differences in that coefficient among species** can be explained by a measured trait. The trait acts on the environmental response, not directly on the occurrence observation.

We have 100 sites to help estimate each species' environmental response, but only ten species whose responses can be compared with their traits. More PCRs do not create more species-level contrasts. More sites can improve the estimated species responses, but they do not remove all uncertainty in the trait relationship.

```{r extract-trait-coefficients, eval=FALSE}
trait_draws <- occJSDM::returnTraitsCoeff(fitmodel)

# Draws for how Trait_1 changes the response to environmental gradient 1.
trait_1_draws <- trait_draws[, "X_psi.EnvCov.1", "Trait_1"]
```

The generating trait matrix is `known_truth$jsdmParams_true$G`. The simulator used unstandardized traits, whereas fitting standardized them. A raw coefficient of -1 therefore does not necessarily appear as -1 on the fitted scale. The comparison below multiplies each generating trait coefficient by that trait's sample standard deviation. The environmental scale already matches.

```{r trait-effects, fig.height=6}
trait_effects <- outputs$coefficients |>
  filter(block == "Trait")

ggplot(trait_effects, aes(x = term)) +
  geom_hline(yintercept = 0, colour = "grey55", linetype = "dashed") +
  geom_pointrange(
    aes(y = estimate, ymin = lower, ymax = upper),
    colour = "#0072B2"
  ) +
  geom_point(aes(y = truth), shape = 4, size = 3, stroke = 1) +
  facet_grid(
    covariate ~ arm,
    labeller = labeller(covariate = environment_labels, arm = fit_labels)
  ) +
  labs(
    x = "Measured species trait",
    y = "Change in environmental coefficient per trait standard deviation",
    caption = "Black cross: generating trait effect on the fitted scale. Blue: posterior mean and 95% interval."
  )
```

Three of the four generating trait effects are nonzero. Only one of the four intervals excludes zero in either fit. That is a genuine limitation of recovery in this example. It is not evidence that the simulation omitted trait effects, and removing observation error does not make it disappear.

### A real cancellation inside this simulated community

The model allows a species' environmental response to combine three contributions: its measured traits, unmeasured species traits, and remaining species differences. These are distinct from the **site** factors used to describe unmeasured conditions at sites.

In these ten species, Trait_1 happens to correlate with the unmeasured species trait (`r round(outputs$trait_hidden_correlation["Trait_1", 1], 2)`). The generator drew these traits independently; a small realized sample can still contain a substantial correlation. For environmental gradient 1, the resulting contributions oppose one another.

The following is an **oracle diagnostic**, possible only because we have the simulation truth. Regress the known species environmental coefficients on both measured traits, then repeat the regression for each generating contribution. Because these regressions use the same predictors, the contributions add to the observed relationship among these ten species.

```{r oracle-trait-regression}
species_comparison <- lesson$input$sim$data_list$traits |>
  as_tibble() |>
  mutate(across(everything(), ~ as.numeric(scale(.x)))) |>
  mutate(true_response = known_truth$jsdmParams_true$B[1, ])

oracle_regression <- lm(
  true_response ~ Trait_1 + Trait_2,
  data = species_comparison
)

enframe(coef(oracle_regression), name = "term", value = "relationship") |>
  knitr::kable(digits = 3)
```

The coefficient of Trait_1 here describes the net relationship in the ten generated species, after adjusting for Trait_2. It is not the isolated generating effect of Trait_1. The decomposition below shows the difference.

```{r trait-cancellation, fig.height=4.5}
cancellation <- outputs$trait_components |>
  filter(trait == "Trait_1", covariate == "X_psi.EnvCov.1") |>
  select(component, value)

cancellation <- bind_rows(
  cancellation,
  tibble(
    component = "Net relationship among these ten species",
    value = sum(cancellation$value)
  )
)

cancellation |>
  mutate(component = factor(component, levels = rev(component))) |>
  ggplot(aes(x = value, y = component)) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col(fill = "#595959", width = 0.6) +
  geom_text(aes(label = sprintf("%+.2f", value)),
            hjust = ifelse(cancellation$value >= 0, -0.15, 1.15)) +
  scale_x_continuous(expand = expansion(mult = 0.18)) +
  labs(
    x = "Contribution to the Trait_1 relationship with environmental response 1",
    y = NULL,
    caption = "All four bars come from known simulation components; none is an estimated occJSDM effect."
  )
```

The negative generating effect is about `r round(cancellation$value[1], 2)`, but the unmeasured-trait contribution is about `r round(cancellation$value[2], 2)`. After adding the remaining differences, the net relationship among the ten species is only `r round(tail(cancellation$value, 1), 2)`. The model must disentangle these overlapping sources using imperfectly estimated species responses. This explains why simply knowing that we put a nonzero effect into the simulator is not a guarantee that its interval will exclude zero.

This is a diagnosis of **this simulation**, not proof that all weak trait results have this cause. It also does not establish the cause of an older fit whose generating values were not saved. A larger, replicated species-sample experiment would be needed to measure how much additional species information helps.

## Residual species associations: did we recover what was put in?

The residual correlation describes the model's shared hidden site component. It is not the raw correlation of PCR detections, and it does not establish a biological interaction between species. Even after supplying every measured environmental covariate, this simulation still contains the deliberately generated hidden site factors.

```{r extract-correlations, eval=FALSE}
residual_correlations <- occJSDM::returnResidualCorrelationMatrix(fitmodel)

# First dimension: lower limit, median, upper limit.
median_correlations <- residual_correlations[2, , ]
```

```{r correlation-truth, fig.height=5}
correlation_plot <- outputs$correlations |>
  filter(arm == "default") |>
  select(species1, species2, Truth = truth, Estimate = estimate) |>
  pivot_longer(c(Truth, Estimate), names_to = "result", values_to = "correlation") |>
  mutate(
    species1 = factor(species1, levels = species_order),
    species2 = factor(species2, levels = rev(species_order)),
    result = factor(result, levels = c("Truth", "Estimate"))
  )

ggplot(correlation_plot, aes(x = species1, y = species2, fill = correlation)) +
  geom_tile() +
  facet_wrap(~ result) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    limits = c(-1, 1), na.value = "grey65"
  ) +
  coord_equal() +
  labs(x = NULL, y = NULL, fill = "Correlation",
       caption = "Estimate: posterior median. Grey truth cells have undefined correlation, not correlation zero.") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

OTU_4 was generated with zero loading on both hidden site factors. Its hidden contribution has no variation, so its correlation with another hidden contribution is mathematically undefined. We show those true cells in grey and exclude them from numerical correlation-error summaries. Painting them white would incorrectly claim their true correlation was zero.

Compare the pattern and magnitude, not just whether an interval crosses zero. An apparently strong estimated correlation can still be far from the generating value.

## Ordination: compare the combined effect before naming the axes

`returnOrdinationScores()` returns hidden site scores; `returnFactorLoadings()` returns species loadings. Multiplying a site's scores by a species' loadings gives their combined contribution to its occurrence predictor. Rotating both sets of axes, or reversing their signs together, can leave that contribution unchanged. Unaligned true and fitted axes are therefore a misleading accuracy comparison.

```{r extract-ordination, eval=FALSE}
site_scores <- occJSDM::returnOrdinationScores(fitmodel)
species_loadings <- occJSDM::returnFactorLoadings(fitmodel)

occJSDM::plotBiplot(fitmodel)
```

For a truth check, the exporter multiplies scores by loadings **within each posterior draw**, then averages the products. Multiplying the mean scores by the mean loadings would not give the same answer.

```{r ordination-contribution, fig.height=4.5}
ggplot(outputs$residual, aes(x = truth, y = estimate)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(alpha = 0.25, size = 1, colour = "#0072B2") +
  facet_wrap(~ arm, labeller = labeller(arm = fit_labels)) +
  coord_equal() +
  labs(
    x = "True combined hidden site contribution (log-odds)",
    y = "Estimated hidden site contribution\n(posterior mean, log-odds)",
    caption = "Each point is a species at a surveyed site. This comparison is invariant to factor rotation."
  )
```

The diagonal is exact recovery. Points closer to zero than their true values show underestimation of the hidden contribution's magnitude. These are fitted-site results: the observations at a site helped estimate its hidden scores.

## Variation partitioning: an allocation within the model

The package divides variation among environmental, spatial and residual-factor components. In this non-spatial example the spatial fraction is zero, so showing the environmental fraction is enough: the residual fraction is one minus it.

The package calls the residual fraction `Biotic`. That label does not mean it has identified biotic interactions. Here the hidden site factors were simulated directly, without simulating ecological interactions.

```{r extract-variation, eval=FALSE}
variation_table <- occJSDM::returnVariancePartitioning(fitmodel)
```

```{r variation-truth, fig.height=4.5}
outputs$variation |>
  filter(component == "Environmental") |>
  ggplot(aes(x = truth, y = estimate, label = species)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_point(colour = "#0072B2", size = 2) +
  geom_text(
    aes(hjust = if_else(truth > 0.9, 1.1, 0.5)),
    nudge_y = 0.025, size = 2.8, check_overlap = TRUE
  ) +
  facet_wrap(~ arm, labeller = labeller(arm = fit_labels)) +
  scale_x_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1)) +
  coord_equal() +
  labs(x = "True environmental fraction", y = "Estimated environmental fraction",
       caption = "Matching package definition on the same sites. Estimates are posterior means.")
```

The true fractions use the same package definition as the estimates: changes in the standard deviation of probabilities when model components are combined, with negative increments truncated and contributions normalized. They are not a generic fraction of raw detection variance explained. That definition matters when comparing this plot with another package's variation partitioning.

For OTU_4, the true environmental fraction is 100% because its hidden site contribution is zero. The estimate does not automatically discover that fact. Other points above the diagonal allocate too much to the environment under this definition; points below it allocate too little. Good-looking occupancy predictions do not guarantee correct attribution among components.

Why does the PCR fit still allocate variation to hidden site factors when their average contributions in the preceding figure are so close to zero? Each posterior draw can contain substantial positive and negative contributions. If the draws disagree about a site's contribution, those values can cancel when averaged. The partition is instead calculated within each draw and then averaged. A contribution whose average is near zero can therefore still account for variation within individual draws.

## Collection effects and detection probabilities

Collection covariates predict whether DNA enters a field sample, conditional on the species being present at the site. They are separate from the environmental covariates predicting the species' distribution.

```{r collection-effects, fig.height=4}
outputs$collection |>
  filter(covariate == "X_theta") |>
  mutate(species = factor(term, levels = species_order)) |>
  ggplot(aes(x = species)) +
  geom_hline(yintercept = 0, colour = "grey55", linetype = "dashed") +
  geom_pointrange(aes(y = estimate, ymin = lower, ymax = upper), colour = "#0072B2") +
  geom_point(aes(y = truth), shape = 4, size = 3) +
  labs(x = "Species", y = "Collection effect\n(per covariate standard deviation)",
       caption = "Black cross: truth. Blue: posterior mean and 95% interval. Compare on the same standardized scale.") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Use `returnCollectionCovariates()` to extract these draws. [Lesson 1](`r lesson_link("occJSDM-first-lesson")`) compares true and estimated PCR detection, laboratory false-positive and field-contamination probabilities, and shows why their values alone cannot classify every positive detection correctly. Its truth accounts for whether simulated reads actually pass the fitted threshold.

### Would more field samples or PCR replicates help detection?

Consider a deliberately restricted question: **if all ten species occupy a site, how many do we expect to detect truly at least once?** Hold collection conditions at their mean, use both primers, and count only detections arising from collected DNA. False-positive detections are excluded from this calculation. This is neither landscape richness nor the number of species an analysis will correctly infer to be present.

For one species, let `theta` be collection probability and `p1`, `p2` the two primer detection probabilities. With `K` PCRs per primer, the probability of missing collected DNA in every PCR is `(1 - p1)^K * (1 - p2)^K`. Combine this with collection failure, then with `M` independent field samples:

```{r expected-detections-formula, eval=FALSE}
missed_in_pcr <- (1 - p1)^K * (1 - p2)^K

detected_in_one_sample <- theta * (1 - missed_in_pcr)

detected_in_any_sample <- 1 - (1 - detected_in_one_sample)^M

expected_species_detected <- sum(detected_in_any_sample)
```

The exporter applies that formula to the known probabilities and to every posterior draw. The ribbons below describe uncertainty in the **expected number**, not the wider variation in the number a single survey might actually detect. The package's `plotCumulativeSpeciesDetections()` instead simulates survey outcomes; do not equate its interval with this interval for an expectation.

```{r detection-effort, fig.height=5}
outputs$detection_effort |>
  mutate(samples = factor(M)) |>
  ggplot(aes(x = K)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#0072B2", alpha = 0.2) +
  geom_line(aes(y = estimate), colour = "#0072B2", linewidth = 0.8) +
  geom_line(aes(y = truth), linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ samples, ncol = 2, labeller = label_both) +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(limits = c(0, 10)) +
  labs(
    x = "PCR replicates per primer (two primers)",
    y = "Expected species detected out of ten",
    caption = "samples = field samples per site. Black dashed: truth. Blue: posterior mean and 95% interval."
  )
```

Repeated PCRs cannot recover DNA that never entered the field sample. That is why the one-sample curve levels off even when PCR replication increases. Another independent field sample gives another opportunity to collect the species' DNA. These curves hold the fitted parameter distribution fixed; they do not measure how collecting more data would improve a refitted model.

## Distinguish fitted probabilities, occupancy states and new-site predictions

`computePredictiveOccupancyProbs()` returns the fitted ecological probability at surveyed sites. Despite its name, the fitted site factors were learned using the observations. Calling it a prediction based only on measured environmental covariates is incomplete.

`computeConditionalOccupancyProbs()` summarizes the model's belief that the species actually occupied the surveyed site, accounting for the observation process. Its appropriate simulation check is the realized 0/1 state, not the generating probability. [Lesson 1](`r lesson_link("occJSDM-first-lesson")`) puts these quantities alongside the actual simulated detection cases, with maps in Lesson 0.

`returnLatentPresences()` and `plotLatentPresences()` collect those fitted quantities by site, sample and primer. They are useful diagnostic displays, but the old vignette's screenshot came from another simulation. Use the matched examples in Lesson 1 instead.

Genuine prediction at an unsurveyed site requires keeping its observations out of fitting and averaging appropriately over its unknown conditions. Reusing the fitting sites' covariates is not an independent prediction test. A dedicated new-site example and the spatial outputs belong with the planned prediction and spatial lessons; this lesson makes no held-out or spatial-accuracy claim.

## Check computation as well as ecological recovery

An interval can be wide because the data contain little information, because model components are hard to separate, or because the sampler has not adequately explored its posterior. The figures alone do not distinguish those causes. Check the numerical diagnostics before interpreting uncertainty.

Think of each MCMC chain as a separate exploration of the parameter values that could explain the observations. After discarding the initial settling-in period, the chains should explore similar distributions. They need not take the same path or return the true value on every iteration. Our practical sequence is: obtain the diagnostics, identify individual parameters needing attention, inspect their traces, and decide whether further computation or investigation is needed.

### Get the diagnostics for your own fit

Use the `fitmodel` returned by `runOccJSDM()`. If you are reproducing our saved example instead, load its complete fit first:

```{r extract-full-fit-diagnostics, eval=FALSE}
saved_fit <- readRDS("/path/to/full-fits/default-fit.rds")
fitmodel <- saved_fit$fit

parameter_diagnostics <- occJSDM::returnConvergenceDiagnostics(fitmodel)

parameter_diagnostics |>
  select(param, label1, label2, rhat, ess)
```

There is one row per parameter, such as the collection effect for one species or the detection rate for one species and primer. `idx1` and `idx2` are positions in the saved arrays; `label1` and `label2` identify what those positions mean. A placeholder `"1"` in `label2` for a species-only parameter is not another species or primer.

| `param` | Ecological meaning | `label1` | `label2` |
|---|---|---|---|
| `beta0_psi` | Baseline occupancy on the log-odds scale | Species | Placeholder |
| `beta_psi` | Environmental effect on occupancy | Environmental covariate | Species |
| `beta_theta` | Collection intercept or covariate effect | `(Intercept)` or collection covariate | Species |
| `p` | Positive PCR probability when DNA is in the sample | Primer | Species |
| `q` | Positive PCR probability when DNA is absent from the sample | Primer | Species |
| `theta0` | Field-contamination probability when the species is absent from the site | Species | Placeholder |

The table also contains `mean`, `sd`, `q2.5` and `q97.5`: posterior summaries of the corresponding parameter. They are distinct from the numerical checks `rhat` and `ess`. The public function covers these six parameter blocks, **not every quantity in the model**. In particular, it omits the trait coefficients and the individual latent-factor arrays.

### Find the parameters that need attention

The following code uses the same public diagnostic table saved with each teaching fit. When working with your own model, replace the first assignment with the call above.

```{r flag-default-diagnostics}
parameter_diagnostics <- as_tibble(lesson$diagnostics$default)

flag_parameters <- function(diagnostics) {
  diagnostics |>
    mutate(
      unavailable = !is.finite(rhat) | !is.finite(ess),
      high_rhat = rhat > 1.01,
      low_ess = ess < 400
    ) |>
    filter(unavailable | high_rhat | low_ess) |>
    arrange(desc(unavailable), desc(rhat), ess)
}

flagged_default <- flag_parameters(parameter_diagnostics)

nrow(flagged_default)
```

For this default-prior fit, `r nrow(flagged_default)` rows need attention under this screen. An empty list means that the listed parameters pass these checks; it is not a certificate that all quantities have converged or that the ecological estimates are accurate. Missing diagnostics are flagged too. A missing Rhat can arise from having only one chain or a chain that never moved, so silently dropping missing values would hide a possible problem.

Now apply exactly the same screen to the **longer alternative-prior fit** from Lesson 1. That fit uses weaker low-contamination assumptions and has already been extended to four chains with 12,000 retained iterations each:

```{r flag-alternative-diagnostics}
flagged_alternative <- as_tibble(lesson$diagnostics$alternative) |>
  flag_parameters()

flagged_alternative |>
  select(param, label1, label2, rhat, ess, unavailable, high_rhat, low_ess) |>
  knitr::kable(digits = c(0, 0, 0, 4, 0, 0, 0, 0))
```

Here the rows identify the actual parameter to investigate. For example, `theta0` for OTU_6 has a low effective sample size even though its Rhat is just below 1.01. Looking at Rhat alone would miss that warning. The collection intercept for OTU_6 and the first environmental slope for OTU_1 are flagged for Rhat instead.

These are screening rules, not sharp boundaries between trustworthy and untrustworthy results. In this package revision, `returnConvergenceDiagnostics()` uses the classical `coda` Rhat and effective sample size calculations. The newer `posterior` calculations below use rank-normalized split-chain Rhat and distinguish bulk from tail ESS. Their numbers can differ; do not relabel the public `ess` column as bulk or tail ESS. The [Stan diagnostics guide](https://mc-stan.org/learn-stan/diagnostics-warnings.html) explains the newer diagnostics and the commonly used 1.01 Rhat screen. Our legacy ESS screen of 400 is a prompt to inspect precision, not a guarantee about every posterior summary.

### Read a traceplot for a collection covariate

A traceplot shows the sampled value against iteration, with a colour for each chain. First locate the collection covariate by name. The array has dimensions `[covariate, species, iteration, chain]`; `drop = FALSE` retains those four dimensions after selecting one covariate.

```{r extract-collection-trace, eval=FALSE}
collection_name <- "X_theta"
collection_index <- match(collection_name, colnames(fitmodel$X_theta))

stopifnot(!is.na(collection_index))

collection_draws <- fitmodel$results_output$beta_theta_output[
  collection_index, , , , drop = FALSE
]

occJSDM::plotTraceplot(
  collection_draws,
  param_name = "Collection effect (log-odds)",
  dimnames1 = collection_name,
  dimnames2 = fitmodel$infos$speciesNames
)
```

That call plots every species. For this page, the small saved bundle retains all draws and all four chains for OTU_1 and OTU_6 so the panels remain readable. These are excerpts of the full fit, not newly fitted models. We add the known collection effects on the **fitted standardized-covariate scale**, using the same scale conversion as the coefficient figure above.

```{r collection-trace, fig.height=5}
collection_trace <- diagnostic_examples$traces$collection

occJSDM::plotTraceplot(
  collection_trace$draws,
  param_name = "Collection effect (log-odds)",
  dimnames1 = collection_trace$label1,
  dimnames2 = collection_trace$label2
) +
  geom_hline(
    data = collection_trace$truth,
    aes(yintercept = truth), colour = "black", linetype = "dashed"
  ) +
  facet_grid(label2 ~ chain, scales = "free_y") +
  labs(
    x = "Saved draw after burn-in", colour = "Chain",
    caption = "Columns: chains 1 to 4. Rows: species. Dashed line: generating effect. All 6,000 saved draws per chain are shown."
  ) +
  theme(legend.position = "none")
```

Compare the height and spread of the traces between columns for the same species. Persistent separation between chains, a continuing upward or downward drift, or long periods stuck in a narrow region would warrant investigation. Jagged movement and a wide vertical spread are not by themselves failures: they may reflect a broad posterior distribution.

The black line asks a different question: **where is the generating effect relative to the sampled values?** Agreement between chains checks whether the calculation is stable; proximity to the line checks recovery in this simulation. Chains can agree while being centred away from truth, or agree on a wide interval containing truth. Do not change the sampler merely to force its traces onto the black line.

### Extract a primer and inspect its detection rate

The same method applies to `p` and `q`, but their first dimension is primer rather than covariate. Primer identifiers may be stored as numbers, so convert their labels to character before matching.

```{r extract-primer-trace, eval=FALSE}
primer_name <- "1"
primer_index <- match(primer_name, as.character(fitmodel$infos$primerNames))

stopifnot(!is.na(primer_index))

primer_draws <- fitmodel$results_output$p_output[
  primer_index, , , , drop = FALSE
]

occJSDM::plotTraceplot(
  primer_draws,
  param_name = "PCR detection probability",
  dimnames1 = primer_name,
  dimnames2 = fitmodel$infos$speciesNames
)
```

```{r primer-trace, fig.height=5}
primer_trace <- diagnostic_examples$traces$detection

occJSDM::plotTraceplot(
  primer_trace$draws,
  param_name = "PCR detection probability",
  dimnames1 = primer_trace$label1,
  dimnames2 = primer_trace$label2
) +
  geom_hline(
    data = primer_trace$truth,
    aes(yintercept = truth), colour = "black", linetype = "dashed"
  ) +
  facet_grid(label2 ~ chain, scales = "free_y") +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "Saved draw after burn-in",
    caption = "Primer 1, default-prior fit. Columns: chains. Rows: species. Truth accounts for the fitted read threshold."
  ) +
  theme(legend.position = "none")
```

These true probabilities include the chance that a simulated detection event produces enough reads to count as positive. Using just the simulator's event probability would put the reference line at the wrong target. The collection and PCR trace examples use the same two species, so neither panel was selected for especially good recovery.

### Follow up an actual diagnostic warning

For a species-only parameter such as `theta0`, the saved array has three dimensions: `[species, iteration, chain]`. Do not apply the four-index expression above to it. This example selects the field-contamination parameter with the lowest public ESS in the longer alternative-prior fit, which is OTU_6. Selection uses diagnostics, not distance from truth.

```{r extract-field-trace, eval=FALSE}
alternative_fit <- readRDS("/path/to/full-fits/alternative-long-fit.rds")$fit
species_index <- match("OTU_6", alternative_fit$infos$speciesNames)

stopifnot(!is.na(species_index))

field_draws <- alternative_fit$results_output$theta0_output[
  species_index, , , drop = FALSE
]

occJSDM::plotTraceplot(
  field_draws,
  param_name = "Field-contamination probability",
  dimnames1 = "OTU_6"
)
```

```{r flagged-field-trace, fig.height=5.5}
field_trace <- diagnostic_examples$traces$field_contamination

occJSDM::plotTraceplot(
  field_trace$draws,
  param_name = "Field-contamination probability",
  dimnames1 = field_trace$label1
) +
  geom_hline(
    data = field_trace$truth,
    aes(yintercept = truth), colour = "black", linetype = "dashed"
  ) +
  facet_wrap(~ chain, ncol = 1) +
  scale_x_continuous(breaks = seq(0, 12000, by = 3000)) +
  scale_y_continuous(labels = scales::label_percent()) +
  labs(
    x = "Saved draw after burn-in",
    caption = "OTU_6, longer alternative-prior fit. Panels: chains. Dashed line: true field-contamination probability."
  ) +
  theme(legend.position = "none")
```

There are 48,000 retained draws here, but their dependence means they carry much less independent information. The public ESS for this parameter is about `r round(flagged_alternative$ess[flagged_alternative$param == "theta0"])`. That is a numerical limitation on posterior summaries, not a count of field samples and not a claim that only that many iterations were run. This fit should not be described as having cleared every diagnostic simply because it was run for longer.

The traces make this slow movement visible: each chain spends stretches at relatively high or low contamination rates. The true rate is `r scales::percent(field_trace$truth$truth, accuracy = 0.1)`, near the bottom of the panels, but the fit also gives substantial weight to much higher rates. The example therefore raises two separate concerns: how precisely the sampler has estimated its posterior summaries, and how well that posterior recovers the ecological truth.

### Where to find other parameter draws

Start from `fitmodel$results_output`. Keep the iteration and chain dimensions separate when calculating diagnostics; pooling chains into one vector destroys the information Rhat needs.

| Parameter | Array within `results_output` | Dimensions before selecting a parameter |
|---|---|---|
| Occupancy intercept | `jsdm_output$B0_output` | Species, iteration, chain |
| Environmental effect | `jsdm_output$B_output` | Covariate, species, iteration, chain |
| Measured trait effect | `jsdm_output$G_output` | Trait, environmental covariate, iteration, chain |
| Collection intercept/effect | `beta_theta_output` | Covariate, species, iteration, chain |
| PCR detection/false-positive rate | `p_output`, `q_output` | Primer, species, iteration, chain |
| Field-contamination rate | `theta0_output` | Species, iteration, chain |

For example, direct newer diagnostics for one environmental coefficient use an **iteration-by-chain matrix**. This optional code requires the `posterior` package, also used by our offline verification scripts:

```{r direct-coefficient-diagnostics, eval=FALSE}
environment_index <- match("X_psi.EnvCov.1", colnames(fitmodel$X_psi))
species_index <- match("OTU_1", fitmodel$infos$speciesNames)

stopifnot(!anyNA(c(environment_index, species_index)))

coefficient_draws <- fitmodel$results_output$jsdm_output$B_output[
  environment_index, species_index, ,
]

tibble(
  Rhat = posterior::rhat(coefficient_draws),
  bulk_ESS = posterior::ess_bulk(coefficient_draws),
  tail_ESS = posterior::ess_tail(coefficient_draws)
)
```

### Summaries and what to do next

```{r fitting-diagnostics}
diagnostic_summary <- bind_rows(
  as_tibble(lesson$diagnostics$perfect) |> mutate(arm = "perfect"),
  as_tibble(lesson$diagnostics$default) |> mutate(arm = "default")
) |>
  group_by(arm) |>
  summarise(
    parameters = n(),
    unavailable = sum(!is.finite(rhat) | !is.finite(ess)),
    maximum_Rhat = max(rhat[is.finite(rhat)]),
    minimum_ESS = min(ess[is.finite(ess)]),
    .groups = "drop"
  ) |>
  mutate(fit = unname(fit_labels[arm])) |>
  select(fit, everything(), -arm)

knitr::kable(diagnostic_summary, digits = 3)
```

These are the saved parameter diagnostics, not a certification of unbiased inference. Good chain agreement does not resolve limited species information or separate overlapping ecological explanations. Use `returnConvergenceDiagnostics()` and `plotTraceplot()` on the full fit for parameter-level checks; a traceplot can include the corresponding true value as a reference.

The public diagnostics table does not include the trait matrix `G`. Because trait recovery is a main concern here, the exporter also calculates diagnostics directly from the environmental and trait coefficient draws, retaining their chain identities:

```{r coefficient-diagnostics}
outputs$coefficients |>
  group_by(arm, block) |>
  summarise(
    maximum_Rhat = max(rhat),
    minimum_bulk_ESS = min(ess_bulk),
    minimum_tail_ESS = min(ess_tail),
    .groups = "drop"
  ) |>
  mutate(fit = unname(fit_labels[arm])) |>
  select(fit, block, maximum_Rhat, minimum_bulk_ESS, minimum_tail_ESS) |>
  knitr::kable(digits = c(0, 0, 3, 0, 0),
               caption = "Direct checks of environmental and trait coefficient sampling.")
```

Rhat close to one indicates agreement among chains. Effective sample size describes how much independent information remains in the correlated MCMC draws; the tail calculation is relevant to interval endpoints. These diagnostics check computation, not whether the available ecological data can identify the generating trait effect. See the [`posterior` diagnostics documentation](https://mc-stan.org/posterior/reference/diagnostics.html) for the definitions used in this coefficient table.

When a parameter is flagged, inspect its trace and consider which ecological conclusions depend on it. If the chains explore similar distributions but move slowly, more iterations may improve precision. If they remain separated or drift, investigate the model, priors, data information and starting values rather than assuming a longer run will necessarily solve the problem. Recheck both diagnostics and the stability of the reported estimates after any change.

Keep `nthin = 1` unless storage is the limiting concern. Discarding additional draws does not make the sampler explore better or repair chains trapped in different regions. Never discard selected chains or tune priors just to make the examples pass a diagnostic threshold. In a real dataset the black truth lines are unavailable, so convergence checks, model checks and ecological judgment each have a separate role.

WAIC is another diagnostic quantity with no generating parameter to overlay. `extractWAIC()` can support comparison of models fitted to the same observations, subject to its assumptions. The perfect-observation and PCR fits here have different response data, so their WAIC values must not be compared as if they were competing models of one dataset. We have not fitted a model-selection example in this lesson.

## Reproduce the extraction or find a function

The compact files retain source and fit hashes. Full MCMC fits are kept outside the package because they are much larger. To regenerate them, follow `dev/simstudy/vignette-lesson/README.md` and run the existing simulation/fitting commands with the recorded source. Then export this lesson's additional summaries:

```{bash, eval=FALSE}
Rscript dev/simstudy/vignette-lesson/summarise_outputs.R /path/to/full-fits
Rscript dev/simstudy/vignette-lesson/verify_outputs.R /path/to/full-fits
Rscript dev/simstudy/vignette-lesson/summarise_diagnostics.R /path/to/full-fits
Rscript dev/simstudy/vignette-lesson/verify_diagnostics.R /path/to/full-fits
```

In an R session with those full fits available:

```{r load-full-fit, eval=FALSE}
saved_fit <- readRDS("/path/to/full-fits/default-fit.rds")
fitmodel <- saved_fit$fit

known_truth <- lesson$input$sim$true_params
```

The exporter preserves all posterior draws for its summaries. It does not rerun MCMC or choose a different community because an effect was not recovered. Student-facing figures use tidy tables so the plotting code remains readable; the export and verification scripts document and check the array calculations behind them.

| Ecological question | Useful functions | Truth check in the lessons |
|---|---|---|
| How do I prepare and fit data? | `simulateOccJSDMData()`, `runOccJSDM()` | Lessons 0 and 1 |
| How does each species respond to the environment? | `returnOccupancyCovariates()`, `plotOccupancyCovariates()`, `returnOccupancyGradient()`, `plotOccupancyGradient()`, `plotCovariateEffect()` | Coefficients and response profiles above |
| What is baseline occupancy? | `returnOccupancyRates()`, `plotOccupancyRates()` | Baseline table above |
| Do traits explain species responses? | `returnTraitsCoeff()`, `plotTraitsCoefficients()` | Trait estimates and cancellation diagnostic above |
| Which species share unmeasured site responses? | `returnResidualCorrelationMatrix()`, `plotResidualCorrelationMatrix()` | Matched correlation matrices above |
| What do ordination axes represent? | `returnOrdinationScores()`, `returnFactorLoadings()`, `plotOrdinationScores()`, `plotFactorLoadings()`, `plotBiplot()` | Rotation-invariant combined contribution above |
| How is variation allocated? | `returnVariancePartitioning()`, `plotVariancePartitioning()` | Matching true and fitted fractions above |
| What affects collection? | `returnCollectionCovariates()`, `plotCollectionCovariates()`, `plotCollectionRates()` | Collection effects above; observation process in Lesson 1 |
| What about PCR failures and contamination? | `plotDetectionRates()`, `plotStage1FPRates()`, `plotStage2FPRates()` | Rate recovery and actual cases in Lesson 1 |
| How does sampling effort affect detection? | `plotCumulativeSpeciesDetections()` | Analytic expectation above, with its interval distinction |
| What happened at a particular site/sample? | `computeConditionalOccupancyProbs()`, `computePredictiveOccupancyProbs()`, `returnLatentPresences()`, `plotLatentPresences()` | Matched probabilities, states and observations in Lesson 1 |
| Can I trust the computation? | `returnConvergenceDiagnostics()`, `plotTraceplot()`, `extractWAIC()` | Diagnostics above; these have no single simulated true value |
| What about space or unsurveyed sites? | `predictNewSites()` and spatial model outputs | Dedicated examples still needed; not validated by this lesson |
