---
title: "simulateOccPlusData"
output: 
  rmarkdown::html_vignette:
    fig.width: 7
    fig.height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{simulateOccPlusData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  comment = "#>"
)

options(warning = -1)
```

```{r setup, include = FALSE}
# library(occPlus)
devtools::load_all() # change to library(occPlus) when package finished
library(tidyverse)
```

`simulateOccPlusData()` simulates a full occJSDM-style dataset: species
occupancy driven by environmental covariates, species traits, and
(optionally) a spatially autocorrelated random field, plus a two-stage
eDNA collection/detection process on top of the true occupancy states.

The function takes three lists of settings, plus a `useSpatField`
switch:

- `list_datasettings`: sample-size and dimension settings -- `n`
  (sites), `S` (species), `g` (number of observed trait covariates), `M`
  (visits per site), `P` (primers), `K` (PCR replicates per
  site-visit-primer), `ncov_psi` (occupancy covariates), and
  `ncov_theta` (eDNA collection covariates).
- `list_jsdmParams`: JSDM structural parameters -- `gt` (latent trait
  dimensions), `d` (latent factor dimensions capturing residual species
  co-occurrence), `ds` (spatial latent factor dimensions),
  `sigma_b`/`sigma_bs`/`sigma_ts`/`sigma_h` (variance components), and
  `l_s` (spatial length scale).
- `list_params`: detection-process parameters -- `p` and `q` (detection
  and false-positive amplification rates, `P` x `S` matrices), `theta0`
  and `theta_baseline` (baseline eDNA collection probabilities given
  absence/presence).
- `useSpatField`: if `TRUE`, occupancy incorporates a spatially
  autocorrelated Gaussian process over the simulated site coordinates
  `Xs`; if `FALSE` (the default), `Xs` is still simulated and returned
  but does not influence occupancy.

## Setting up shared simulation parameters

We use the same dimension, trait, and detection-process settings for
both datasets, so that the *only* difference between them is whether the
spatial field is switched on.

```{r}
set.seed(5162)

n <- 100          # sites
M <- rep(2, n)    # visits per site
P <- 2            # primers/markers
S <- 6            # species/OTUs
g <- 3            # observed trait covariates
gt <- 2           # unobserved (latent) trait dimensions
d <- 2            # latent factor dimensions (residual co-occurrence)
ds <- 2           # spatial latent factor dimensions
ncov_psi <- 2     # environmental covariates for occupancy
ncov_theta <- 2   # covariates for eDNA collection probability

N <- sum(M)
K <- rep(3, P * N)  # PCR replicates per site-visit-primer combo

list_datasettings <- list(
  n = n, S = S, g = g, M = M, P = P, K = K,
  ncov_psi = ncov_psi, ncov_theta = ncov_theta
  )

list_jsdmParams <- list(
  gt = gt, d = d, ds = ds,
  sigma_b = 0.5, sigma_bs = 0.5, sigma_ts = 0.5,
  sigma_h = 1, l_s = 0.3
  )

list_params <- list(
  p = matrix(runif(P * S, 0.6, 0.95), P, S),   # detection prob given amplicon present
  q = matrix(runif(P * S, 0.01, 0.05), P, S),  # false-positive amplification rate
  theta0 = runif(S, 0.02, 0.1),                # baseline collection prob given absent
  theta_baseline = runif(S, 0.3, 0.7)          # baseline collection prob given present
  )
```

## Simulating without spatial autocorrelation

```{r}
sim_nospat <- simulateOccPlusData(
  list_datasettings, list_params, list_jsdmParams,
  useSpatField = FALSE
  )

str(sim_nospat$data_list, max.level = 1)
```

`data_list` always contains the trait matrix `Tr` (`S` x `g`) and site
coordinates `Xs` (`n` x `2`), regardless of `useSpatField`. With
`useSpatField = FALSE`, `Xs` is generated but does not enter the
occupancy process.

## Simulating with spatial autocorrelation

Using the same settings, we now switch on the spatial field. The spatial
component uses `ds` latent spatial factors and length scale `l_s` to
build a Gaussian-process-correlated random field over `Xs`, which is
added to each species' linear predictor for occupancy.

```{r}
sim_spat <- simulateOccPlusData(
  list_datasettings, list_params, list_jsdmParams,
  useSpatField = TRUE
  )

str(sim_spat$data_list, max.level = 1)
```

## Checking how much variance is spatial

Each simulation's `true_params$jsdmParams_true$varPart` decomposes the
linear predictor's variance per species into Environmental, Spatial, and
Biotic components (plus a Total row of residual/intercept variance).
This is a quick way to check that turning on `useSpatField` actually
contributes spatial variance.

```{r}
sim_nospat$true_params$jsdmParams_true$varPart
sim_spat$true_params$jsdmParams_true$varPart
```

As expected, the "Spatial" column is (numerically) negligible when
`useSpatField = FALSE`, and becomes a non-trivial share of variance --
though it varies considerably across species -- when
`useSpatField = TRUE`.

## Visualizing simulated occupancy across sites

We can compare the true occupancy states (`z_true`) for a single
species, plotted at its site coordinates, with and without the spatial
field switched on.

```{r}
species_idx <- 5

df_compare <- bind_rows(
  tibble(
    x = sim_nospat$data_list$Xs[, 1],
    y = sim_nospat$data_list$Xs[, 2],
    occupied = factor(sim_nospat$true_params$z_true[, species_idx], levels = c(0, 1)),
    field = "useSpatField = FALSE"
  ),
  tibble(
    x = sim_spat$data_list$Xs[, 1],
    y = sim_spat$data_list$Xs[, 2],
    occupied = factor(sim_spat$true_params$z_true[, species_idx], levels = c(0, 1)),
    field = "useSpatField = TRUE"
  )
)

ggplot(df_compare, aes(x, y, color = occupied)) +
  geom_point(size = 2.5) +
  facet_wrap(~field) +
  labs(
    title = paste0("Simulated true occupancy for species ", species_idx),
    x = "Spatial coordinate 1", y = "Spatial coordinate 2", color = "Occupied (z)"
  )
```

Because occupancy is thresholded to 0/1 and the environmental and biotic
terms often explain more variance than the spatial term (see the
`varPart` tables above), spatial clustering is not always obvious by eye
in the binary occupancy map, even when the underlying field is genuinely
spatially structured. The continuous linear predictor
(`true_params$jsdmParams_true$eta`) is a less noisy way to inspect the
spatial pattern directly:

```{r}
df_eta <- tibble(
  x = sim_spat$data_list$Xs[, 1],
  y = sim_spat$data_list$Xs[, 2],
  eta = sim_spat$true_params$jsdmParams_true$eta[, species_idx]
)

ggplot(df_eta, aes(x, y, color = eta)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  labs(
    title = paste0("Linear predictor (eta) for species ", species_idx, ", useSpatField = TRUE"),
    x = "Spatial coordinate 1", y = "Spatial coordinate 2", color = "eta"
  )
```

To make the spatial signal visually dominant, increase `l_s` (broader,
smoother spatial patches) and/or reduce `sigma_h`/`sigma_b` (less
non-spatial noise from latent factors and environmental covariates).

## How traits shape species-specific occupancy responses

Species' responses to environmental covariates are not simulated
independently: each species' coefficient vector for `X_psi` is built
partly from its trait values `Tr`, via a trait-covariate coefficient
matrix `G` (`g` x `ncov_psi`). Specifically,

$$B = (Tr \, G + A \, C + B_t)^\top,$$

where `A` and `C` represent an *unobserved* trait dimension and its
response, and `Bt` is residual, trait-independent variation. `B`
(`ncov_psi` x `S`) is the matrix of true environmental-covariate
coefficients actually used to simulate occupancy.

```{r}
G_true <- sim_nospat$true_params$jsdmParams_true$G
B_true <- sim_nospat$true_params$jsdmParams_true$B
Tr_true <- sim_nospat$data_list$Tr

G_true   # g x ncov_psi: how each trait shifts each covariate's coefficient
```

`G`'s entries are drawn from `{-1, 0, 1}`, so each trait either has no
effect, a positive effect, or a negative effect on a given covariate's
coefficient. We can isolate the trait-driven component of `B`,
`Tr %*% G`, and compare it to the full coefficient matrix to see how
much of species-to-species variation in the covariate response is
attributable to traits versus the unobserved/residual terms.

```{r}
trait_component <- Tr_true %*% G_true   # S x ncov_psi

df_trait <- tibble(
  species = paste0("OTU_", seq_len(S)),
  trait1 = Tr_true[, 1],
  coef_cov1_full = B_true[1, ],
  coef_cov1_trait_only = trait_component[, 1]
)

df_trait
```

Plotting each species' true coefficient for the first covariate against
its value for the first trait shows the trait-driven relationship,
though the full coefficient (which also includes the unobserved-trait
and residual terms) will scatter around it rather than fall exactly on a
line.

```{r}
ggplot(df_trait, aes(trait1, coef_cov1_full)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "Species' occupancy response to covariate 1 vs. trait 1",
    x = "Trait 1 value", y = "True coefficient for covariate 1 (B)"
  )
```

The sign and magnitude of the slope is governed by `G_true[1, 1]` (trait
1's effect on covariate 1); when it is `0`, species' coefficients vary
with trait 1 only through the unobserved-trait pathway (`A %*% C`) and
residual noise (`Bt`), so no relationship is expected.
